---
title: "COPD"
output: html_document
date: "2026-03-27"
---
# Load in the libraries
```{r}
library(R2jags)
library(coda)
library(ggplot2)
library(GGally)
library(dplyr)
library(loo)
library(runjags)
```
# Data Cleaning
```{r}
load("/Users/yanfei/Desktop/Data.RData")
```
# Model Fitting
# COPD-PM2.5, PM10, O3, Temp
# Pneumonia-PM2.5, PM10, SO2, O3, Temp
# HeartFailure-PM2.5, SO2, NO2, CO, O3, Temp
```{r}
Y <- cbind(
  # log(df$PM2.5)-mean(log(df$PM2.5)),
           # log(df$PM10)-mean(log(df$PM10)),
           # log(df$SO2)-mean(log(df$SO2)),
           log(df$NO2)-mean(log(df$NO2)),
           log(df$CO)-mean(log(df$CO)),
           log(df$O3)-mean(log(df$O3)))
# Filter only the first 258 timepoints
Y=Y[1:258,]
N <- 258
N1 <- 258
```

```{r}
# Number of pollutants included in the outcome Poisson Model
K <- 3
```
# Singular value decomposition Transformation
```{r}
svd_res <- svd(scale(Y, center = TRUE, scale = FALSE))
# Transformation Matrix
Q <- svd_res$v
# Pollution Level in the Transformed Space
Y_rot <- t(Q) %*% t(Y)
Y_rot <- t(Y_rot)  # N x K
# Sin and Cos for explaining the seasonal pattern
n_harmonics <- 1  # Number of harmonics to include
sin_terms <- matrix(NA, nrow = N, ncol = n_harmonics)
cos_terms <- matrix(NA, nrow = N, ncol = n_harmonics)
for (h in 1:n_harmonics) {
  sin_terms[, h] <- sin(2 * pi * h * (1:N) / 52)
  cos_terms[, h] <- cos(2 * pi * h * (1:N) / 52)
}
```
# Prepare Data for the jags model
```{r}
data_jags <- list(
  Y = Y_rot,
  Y_out = data$Neoplasms[2:(N1+1)],
  N = N,
  N1 = N1,
  K = K,
  pi = pi,
  tv = 1:N,
  
  sin_harmonics = sin_terms,
  cos_harmonics = cos_terms,
  n_harmonics = n_harmonics,
  
  lam = 1/52,
  yv = df$Temperature[1:N],
  offset = data$Population[2:(N1+1)],
  Q = Q
)

```
# Define the model
```{r}
model_string_flexible <- "
model {

  # Latent states with flexible seasonal structure
  # ----------------------------------------------------------
  for (k in 1:K) {
    # Base mean with trend and flexible seasonal
    for (t in 1:N) {
      
      mu_base[t,k] <- a[k] + b[k]*t + 
                     inprod(beta_harmonic[k, 1:n_harmonics], sin_harmonics[t, 1:n_harmonics]) + 
                     inprod(beta_harmonic_cos[k, 1:n_harmonics], cos_harmonics[t, 1:n_harmonics])
    }
    
    # AR(1) structure
    adjusted_mu[1,k] <- mu_base[1,k]
    for (t in 2:N) {
      adjusted_mu[t,k] <- mu_base[t,k] + rho[k] * (Z_star[t-1,k] - mu_base[t-1,k])
    }
    
    # Block sampling for latent states
    for (t in 1:N) {
      Z_star[t,k] ~ dnorm(adjusted_mu[t,k], tau_proc[k])
    }
  }
  # ----------------------------------------------------------

  # 3. Measurement model
  # ----------------------------------------------------------
  for (t in 1:N) {
    for (k in 1:K) {
      Y[t,k] ~ dnorm(Z_star[t,k], tau_me[k])
    }
  }
  # ----------------------------------------------------------

  # 4. Back-transformation to original scale
  for (t in 1:N) {
    for (j in 1:K) {
      muY_orig[t,j] <- inprod(Q[j,1:K], Z_star[t,1:K])
    }
  }

  # 5. Outcome model
  # ----------------------------------------------------------
  for (t in 1:N1) {
    log_lambda[t] <- alpha + inprod(beta_out[1:K], muY_orig[t,1:K]) + 
                     gamma * yv[t] + log(offset[t])
    lambda[t] <- exp(log_lambda[t])
    Y_out[t] ~ dpois(lambda[t])
  }
  # ----------------------------------------------------------

  # 6. Hierarchical priors
  # ----------------------------------------------------------
  # Global shrinkage parameters
  tau_beta_harmonic ~ dgamma(0.5, 0.5)
  tau_beta_out ~ dgamma(0.5, 0.5)
  
  for (k in 1:K) {
    # AR parameter
    rho[k] ~ dnorm(0, 1)
    
    # Process and measurement precision
    tau_proc[k] ~ dgamma(0.5, 0.5)
    tau_me[k] ~ dgamma(0.5, 0.5)
    
    # Intercept and linear trend
    a[k] ~ dnorm(0, 0.01)
    b[k] ~ dnorm(0, 0.01)
    
    # Flexible seasonal coefficients
    for (h in 1:n_harmonics) {
      beta_harmonic[k, h] ~ dnorm(0, tau_beta_harmonic)
      beta_harmonic_cos[k, h] ~ dnorm(0, tau_beta_harmonic)
    }
  }

  # 7. Outcome model parameters
  # ----------------------------------------------------------
  # Block prior for outcome coefficients
  beta_out[1:K] ~ dmnorm(mu_beta_out[], prec_beta_out[1:K, 1:K])
  
  for (k in 1:K) {
    mu_beta_out[k] <- 0
  }
  
  # Diagonal precision matrix
  for (i in 1:K) {
    for (j in 1:K) {
      prec_beta_out[i, j] <- ifelse(i == j, tau_beta_out, 0)
    }
  }
  
  # Global parameters
  alpha ~ dnorm(0, 0.01)
  gamma ~ dnorm(0, 0.01)
}
"
```


```{r}

# 7. Set the Initial values of the mcmc parameters
# ----------------------------------------------------------
inits_function_flexible <- function() {
  list(
    Z_star = Y_rot + matrix(rnorm(N*K, 0, 0.1), N, K),
    rho = rnorm(K, 0, 0.3),
    tau_proc = rgamma(K, 1, 1),
    tau_me = rgamma(K, 1, 1),
    
    # Multiple harmonics initialization
    beta_harmonic = matrix(rnorm(K * n_harmonics, 0, 0.05), K, n_harmonics),
    beta_harmonic_cos = matrix(rnorm(K * n_harmonics, 0, 0.05), K, n_harmonics),
    
    # Fourier basis initialization (if using)
    # beta_fourier = matrix(rnorm(K * n_fourier, 0, 0.05), K, n_fourier),
    
    beta_out = rnorm(K, 0, 0.05),
    a = rnorm(K, 0, 0.1),
    b = rnorm(K, 0, 0.01),
    tau_beta_harmonic = rgamma(1, 0.5, 0.5),
    tau_beta_out = rgamma(1, 0.5, 0.5),
    alpha = rnorm(1, 0, 0.1),
    gamma = rnorm(1, 0, 0.1)
  )
}
```
# Run the jags model
```{r}
n_chains <- 3
n_burnin <- 10000
n_post <- 3000
n_adapt <- 10
n_thin <- 1

vars_monitor <- c("beta_out", "gamma", "alpha", "rho", "tau_proc", "tau_me",
                  "beta_harmonic", "beta_harmonic_cos", "lambda", 
                  "muY_orig", "Z_star", "tau_beta_out", "a", "b")

# system.time({
n_rep=1
mu_hat <- numeric(n_rep)
p=K+2
theta_hat = matrix (NA,n_rep, p)
# cover_mat = matrix (NA,n_rep, p)
ci_array <- array(NA, dim = c(n_rep, p, 2))
for (r in 1:n_rep){
  # Y_out_sim <- data$Neoplasms[2:(N1+1)]
  
  data_jags <- list(
    Y = Y_rot,
    Y_out = data$Neoplasms[2:(N1+1)],
    N = N,
    N1 = N1,
    K = K,
    pi = pi,
    tv = 1:N,
    
    sin_harmonics = sin_terms,
    cos_harmonics = cos_terms,
    n_harmonics = n_harmonics,
    
    lam = 1/52,
    yv = df$Temperature[1:N],
    offset = data$Population[2:(N1+1)],
    Q = Q
  )
  
  results <- run.jags(
    model = model_string_flexible,
    data = data_jags,
    monitor = vars_monitor,
    n.chains = n_chains,
    burnin = n_burnin,
    sample = n_post,
    thin = n_thin,
    adapt = n_adapt,
    method = "parallel",
    modules = "glm",
    inits = replicate(n_chains, inits_function_flexible(), simplify = FALSE),
    silent.jags = FALSE
  )
  post <- as.matrix(as.mcmc(results))
  keep <- c(paste0("beta_out[", 1:K, "]"), "gamma", "alpha"
           
           
  )
  
  post_sub <- post[, keep]
  # theta_hat[1, ] <- colMeans(post_sub)
  theta_hat[r, ] <- colMeans(post_sub)
  
  ci <- apply(post_sub, 2, quantile, c(0.025, 0.975))  
  # ci is 2 × p
  
  ci_array[r, , 1] <- ci[1, ]   # lower
  ci_array[r, , 2] <- ci[2, ]   # upper
  
}

```
# Histograms of the posterior samples 
```{r}
par(mfrow=c(1,1))
# true_value=c(beta_true,gamma_true,alpha_true)
for(j in 1:p){
  hist(theta_hat[,j],
       main=colnames(theta_hat)[j],
       xlab=paste0(keep[j]))
  # abline(v=true_value[j], col="red", lwd=2)
}


# par(mfrow=c(1,p))
par(mfrow=c(1,1))
# true_value=c(beta_true,gamma_true,alpha_true)
for(j in 1:p){
  hist(theta_hat[,j],
       main=colnames(theta_hat)[j],
       xlab=paste0(keep[j]))
  # abline(v=true_value[j], col="red", lwd=2)
}

```
# ...
```{r}
# number of parameters
p <- dim(ci_array)[2]

# color palette for replications
n_rep <- dim(ci_array)[1]
rep_colors <- rainbow(n_rep)   # different color per replication

p <- dim(ci_array)[2]

# number of replications
n_rep <- dim(ci_array)[1]

# color palette for replications
rep_colors <- rainbow(n_rep)


# 9. Model comparison function
# ----------------------------------------------------------
compare_seasonal_models <- function() {
  # This function would run different seasonal specifications
  # and compare using DIC or WAIC
  
  models <- c("single_sin_cos", "multiple_harmonics", "fourier_basis", "dynamic")
  dic_values <- numeric(length(models))
  
  # Run each model specification and compute DIC
  # ... (implementation depends on which option you choose)
  
  return(data.frame(model = models, DIC = dic_values))
}
```
#...
```{r}
# 10. Extract and visualize seasonal patterns
# ----------------------------------------------------------
extract_seasonal <- function(results, n_harmonics, K) {
  # Extract posterior means of seasonal coefficients
  beta_harmonic_mean <- summary(results, vars = "beta_harmonic")$statistics[, "Mean"]
  beta_harmonic_cos_mean <- summary(results, vars = "beta_harmonic_cos")$statistics[, "Mean"]
  
  # Reshape to matrices
  beta_harmonic_mat <- matrix(beta_harmonic_mean, nrow = K, ncol = n_harmonics)
  beta_harmonic_cos_mat <- matrix(beta_harmonic_cos_mean, nrow = K, ncol = n_harmonics)
  
  # Compute seasonal pattern for each pollutant
  t_grid <- 1:52  # One year
  seasonal_patterns <- matrix(NA, nrow = 52, ncol = K)
  
  for (k in 1:K) {
    for (t in 1:52) {
      seasonal_patterns[t, k] <- sum(beta_harmonic_mat[k, ] * sin(2*pi*(1:n_harmonics)*t/52)) +
        sum(beta_harmonic_cos_mat[k, ] * cos(2*pi*(1:n_harmonics)*t/52))
    }
  }
  
  return(seasonal_patterns)
}
# --------------------------------------------------------------------------
# Diagnostics (optional)
# --------------------------------------------------------------------------
# print(summary(results))
# Assuming 'results' is your runjags model output object
sampler_info <- extract(results, what = "samplers")
print(sampler_info)

samples_full <- as.mcmc.list(results)

```
# Traceplot 
```{r}
# --------------------------------------------------------------------------
# 6. Diagnostics and Analysis
# --------------------------------------------------------------------------

## ----Traceplot of the key parameters----
params_to_plot <- c(
  paste0("beta_out[", 1:K, "]"), "gamma", "alpha",
  paste0("tau_me[", 1:K, "]"), paste0("rho[", 1:K, "]"),
  
  paste0("tau_proc[", 1:K, "]"), paste0("tau_me[", 1:K, "]"),
  paste0("beta1[", 1:K, "]"), paste0("beta2[", 1:K, "]"), paste0("tau_beta_out[", 1:K, "]"),
  paste0("a[", 1:K, "]"), paste0("b[", 1:K, "]"),
  paste0("beta_harmonic[", 1:K, ",1]"), paste0("beta_harmonic_cos[", 1:K, ",1]")

)

full_df_all <- data.frame()
for (param in params_to_plot) {
  for (ch in 1:n_chains) {
    # Check if parameter exists in the samples
    if (param %in% colnames(as.matrix(samples_full[[ch]]))) {
      chain_trace <- as.matrix(samples_full[[ch]])[, param]
      iter <- 1:length(chain_trace)
      chain_df <- data.frame(
        iter = iter,
        value = chain_trace,
        chain = factor(ch),
        param = param
      )
      full_df_all <- rbind(full_df_all, chain_df)
    }
  }
}

if (nrow(full_df_all) > 0) {
  p <- ggplot(full_df_all, aes(x = iter, y = value, color = chain)) +
    geom_line(alpha = 0.7) +
    facet_wrap(~param, scales = "free_y") +
    theme_bw() +
    labs(
      title = "Trace Plots with Block Updating (Separated Measurement Model)",
      x = "Iteration",
      y = "Value",
      color = "Chain"
    )
  ggsave("/Users/yanfei/Desktop/param_traceplot_block.png", p, width = 12, height = 8)
} else {
  warning("No trace plots generated - check parameter names")
}

```
# Effective Sample Sizes
```{r}
## ----Effective Sample Size----
# Get all monitored parameters
all_params <- colnames(as.matrix(samples_full[[1]]))
# Filter to parameters of interest for ESS calculation
params_for_ess <- grep("beta_out|gamma|alpha|rho|tau_", all_params, value = TRUE)

if (length(params_for_ess) > 0) {
  ess <- effectiveSize(samples_full[, params_for_ess])
  print("Effective Sample Sizes for Key Parameters:")
  print(sort(ess))
  
  # Calculate minimum ESS
  min_ess <- min(ess)
  cat("\nMinimum ESS:", min_ess, "\n")
  cat("Recommended minimum ESS for reliable inference: 400\n")
  
  if (min_ess < 400) {
    cat("WARNING: Some parameters have ESS below 400. Consider increasing n_post.\n")
  }
} else {
  warning("No parameters found for ESS calculation")
}
```
# Parameter Summaries
```{r}

# Focus on outcome model parameters
params_to_summarize <- c(paste0("beta_out[", 1:K, "]"), "gamma", "alpha", 
                         paste0("rho[", 1:K, "]"),
                         paste0("tau_beta_out"), paste0("tau_me[", 1:K, "]"))

# Extract samples
post_samples_list <- lapply(samples_full, function(chain) {
  chain_mat <- as.matrix(chain)
  # Get only columns that exist
  existing_cols <- intersect(params_to_summarize, colnames(chain_mat))
  if (length(existing_cols) > 0) {
    return(chain_mat[, existing_cols, drop = FALSE])
  } else {
    return(matrix(NA, nrow = nrow(chain_mat), ncol = 0))
  }
})

# Combine chains
post_samples_combined <- do.call(rbind, post_samples_list)

if (ncol(post_samples_combined) > 0) {
  summary_df <- data.frame(
    Parameter = colnames(post_samples_combined),
    Mean = apply(post_samples_combined, 2, mean),
    SD = apply(post_samples_combined, 2, sd),
    `2.5%` = apply(post_samples_combined, 2, quantile, probs = 0.025),
    `50%` = apply(post_samples_combined, 2, quantile, probs = 0.5),
    `97.5%` = apply(post_samples_combined, 2, quantile, probs = 0.975),
    ESS = if (exists("ess")) {
      # Match ESS values to parameters
      sapply(colnames(post_samples_combined), function(p) {
        if (p %in% names(ess)) ess[p] else NA
      })
    } else NA
  )
  
  print("Parameter Summaries with Block Updating:")
  print(summary_df)
  
  # Save to file
  write.csv(summary_df, "/Users/yanfei/Desktop/parameter_summaries_block.csv", row.names = FALSE)
} else {
  warning("No parameters found for summary")
}
```
# WAIC
```{r}

## ----WAIC Calculation----
# Extract lambda samples
lambda_samples_list <- lapply(samples_full, function(chain) {
  chain_mat <- as.matrix(chain)
  lambda_cols <- grep("^lambda\\[", colnames(chain_mat))
  if (length(lambda_cols) > 0) {
    return(chain_mat[, lambda_cols, drop = FALSE])
  } else {
    return(matrix(NA, nrow = nrow(chain_mat), ncol = data_jags$N1))
  }
})

# Combine chains
lambda_post <- do.call(rbind, lambda_samples_list)

if (ncol(lambda_post) == data_jags$N1) {
  # Calculate log-likelihood matrix
  log_lik_matrix <- matrix(NA, nrow = nrow(lambda_post), ncol = data_jags$N1)
  for (t in 1:data_jags$N1) {
    log_lik_matrix[, t] <- dpois(data_jags$Y_out[t], lambda_post[, t], log = TRUE)
  }
  
  # Calculate WAIC
  waic_result <- waic(log_lik_matrix)
  print("WAIC for Block Updating Model:")
  print(waic_result$waic)
  
  # Also calculate LOO if desired
  loo_result <- loo(log_lik_matrix)
  print("LOO for Block Updating Model:")
  print(loo_result$looic)
} else {
  warning("Lambda samples do not match N1 dimension")
}

```
# Scatterplot
```{r}

## ----Pairwise Scatterplots of ALL outcome parameters----
# Define all outcome model parameters we want to include
outcome_all_params <- c(paste0("beta_out[", 1:K, "]"), "alpha", "gamma")

# Extract data
plot_data_list <- list()
for (ch in 1:n_chains) {
  chain_mat <- as.matrix(samples_full[[ch]])
  
  # Get the parameters that exist in this chain
  params_exist <- intersect(outcome_all_params, colnames(chain_mat))
  
  if (length(params_exist) > 0) {
    # Create a subset data frame
    df_subset <- as.data.frame(chain_mat[, params_exist])
    df_subset$Chain <- factor(paste("Chain", ch))
    plot_data_list[[ch]] <- df_subset
  }
}

# Combine all chains
if (length(plot_data_list) > 0) {
  plot_data <- bind_rows(plot_data_list)
  
  # Create cleaner names for display
  names(plot_data) <- gsub("beta_out\\[(\\d+)\\]", "β[\\1]", names(plot_data))
  
  # Remove chain column for the plotting matrix
  params_for_plot <- setdiff(names(plot_data), "Chain")
  
  if (length(params_for_plot) >= 2) {
    # Create the pairs plot
    p <- ggpairs(
      plot_data,
      columns = params_for_plot,
      mapping = aes(color = Chain, alpha = 0.5),
      upper = list(continuous = wrap("points", size = 0.8, alpha = 0.3)),
      lower = list(continuous = wrap("cor", size = 3)),
      diag = list(continuous = wrap("densityDiag", alpha = 0.7))
    ) +
      scale_color_brewer(palette = "Dark2") +
      theme_bw() +
      labs(
        title = "Posterior Correlations: β_out, α, and γ Parameters",
        subtitle = paste("Block updating model |", length(params_for_plot), "parameters")
      )
    
    # Adjust size based on number of parameters
    n_params <- length(params_for_plot)
    plot_width <- min(15, 4 + n_params * 0.7)
    plot_height <- min(15, 4 + n_params * 0.7)
    
    ggsave(
      "/Users/yanfei/Desktop/outcome_all_params_pairwise.png",
      p,
      width = plot_width,
      height = plot_height,
      dpi = 300
    )
    
    cat("✓ Pairwise plot created with parameters:", 
        paste(params_for_plot, collapse = ", "), "\n")
    
    # Additional: Create a version with just beta_out parameters
    if (K > 1) {
      beta_cols <- grep("^β\\[", params_for_plot, value = TRUE)
      if (length(beta_cols) > 1) {
        p_beta_only <- ggpairs(
          plot_data,
          columns = beta_cols,
          mapping = aes(color = Chain, alpha = 0.5),
          upper = list(continuous = wrap("points", size = 1, alpha = 0.4)),
          lower = list(continuous = wrap("cor", size = 4, digits = 2)),
          diag = list(continuous = wrap("densityDiag", alpha = 0.8))
        ) +
          scale_color_brewer(palette = "Set1") +
          theme_bw() +
          labs(
            title = "Posterior Correlations: β_out Parameters Only",
            subtitle = paste(K, "pollutants")
          )
        
        beta_plot_size <- min(12, 3 + K * 0.8)
        ggsave(
          "/Users/yanfei/Desktop/beta_out_only_pairwise.png",
          p_beta_only,
          width = beta_plot_size,
          height = beta_plot_size,
          dpi = 300
        )
        
        cat("✓ Beta-only pairwise plot saved\n")
      }
    }
    
  } else {
    cat("⚠ Not enough parameters for pairwise plot\n")
  }
}



```
# Posterior Sampled Pollution Level Backtransformed
```{r}

## ----Model Fit Visualization: Back-transformed Predictions----
# Plot fitted vs observed for first pollutant
t_plot <- 1:N
j_plot <- 1

# Extract posterior means of muY_orig
mu_df <- data.frame()
for (ch in 1:n_chains) {
  chain_mat <- as.matrix(samples_full[[ch]])
  # Find muY_orig columns for pollutant j_plot
  mu_cols <- grep(paste0("^muY_orig\\[.*,", j_plot, "\\]$"), colnames(chain_mat))
  
  if (length(mu_cols) > 0) {
    # Get time indices from column names
    for (col in colnames(chain_mat)[mu_cols]) {
      # Extract time index from column name
      time_match <- regmatches(col, regexpr("\\[([0-9]+),", col))
      if (length(time_match) > 0) {
        t <- as.numeric(gsub("[^0-9]", "", time_match))
        if (t <= N) {
          mean_val <- mean(chain_mat[, col])
          mu_df <- rbind(mu_df, data.frame(
            time = t,
            value = mean_val,
            group = paste0("Chain ", ch)
          ))
        }
      }
    }
  }
}

if (nrow(mu_df) > 0) {
  # Observed values
  obs_df <- data.frame(
    time = t_plot,
    value = Y[t_plot, j_plot],
    group = "Observed"
  )
  
  plot_df <- bind_rows(mu_df, obs_df)
  
  p <- ggplot(plot_df, aes(x = time, y = value, color = group, linetype = group)) +
    geom_line(size = 1) +
    scale_linetype_manual(values = c(
      "Chain 1" = "solid", "Chain 2" = "solid", "Chain 3" = "solid",
      "Observed" = "dashed"
    )) +
    theme_bw() +
    labs(
      title = "Model Fit: Back-transformed Predictions vs Observed (Pollutant 1)",
      subtitle = "Block Updating with Separated Measurement Model",
      x = "Time",
      y = "Concentration"
    )
  ggsave("/Users/yanfei/Desktop/pollution_fit_block.png", p, width = 10, height = 6)
}
```
# Observed vs Predicted Mortality
```{r}

# Posterior mean of lambda
lambda_post_list <- lapply(samples_full, function(ch) {
  chain_mat <- as.matrix(ch)
  lambda_cols <- grep("^lambda\\[", colnames(chain_mat))
  if (length(lambda_cols) > 0) {
    return(chain_mat[, lambda_cols, drop = FALSE])
  } else {
    return(NULL)
  }
})

# Remove NULL entries
lambda_post_list <- lambda_post_list[!sapply(lambda_post_list, is.null)]

if (length(lambda_post_list) > 0) {
  lambda_post <- do.call(rbind, lambda_post_list)
  lambda_mean <- apply(lambda_post, 2, mean)
  
  # Data frame
  mort_df <- data.frame(
    time = 1:data_jags$N1,
    observed = data_jags$Y_out,
    predicted = lambda_mean
  )
  
  # Plot
  p <- ggplot(mort_df, aes(x = time)) +
    geom_line(aes(y = observed, color = "Observed"), linetype = "dashed", size = 1) +
    geom_line(aes(y = predicted, color = "Predicted"), size = 1) +
    geom_ribbon(aes(ymin = apply(lambda_post, 2, quantile, 0.025),
                    ymax = apply(lambda_post, 2, quantile, 0.975)),
                alpha = 0.2, fill = "red") +
    scale_color_manual(values = c("Observed" = "black", "Predicted" = "red")) +
    theme_bw() +
    labs(
      title = "Observed vs Predicted Mortality (95% CI)",
      subtitle = "Block Updating Model",
      x = "Time",
      y = "Mortality Count",
      color = ""
    )
  ggsave("/Users/yanfei/Desktop/mortality_fit_block.png", p, width = 10, height = 6)
  
  # Calculate prediction metrics
  rmse <- sqrt(mean((mort_df$observed - mort_df$predicted)^2))
  mae <- mean(abs(mort_df$observed - mort_df$predicted))
  corr <- cor(mort_df$observed, mort_df$predicted)
  
  cat("\n=== Mortality Prediction Metrics ===\n")
  cat("RMSE:", round(rmse, 3), "\n")
  cat("MAE:", round(mae, 3), "\n")
  cat("Correlation:", round(corr, 3), "\n")
}

```
# Log-likelihood trace 
```{r}

loglik_df <- data.frame()

for (ch in 1:n_chains) {
  chain_mat <- as.matrix(samples_full[[ch]])
  lambda_cols <- grep("^lambda\\[", colnames(chain_mat))
  
  if (length(lambda_cols) > 0) {
    lambda_mat <- chain_mat[, lambda_cols, drop = FALSE]
    
    # Compute log-likelihood at each iteration
    loglik_vec <- apply(lambda_mat, 1, function(lam_row) {
      sum(dpois(data_jags$Y_out, lam_row, log = TRUE))
    })
    
    loglik_df <- rbind(
      loglik_df,
      data.frame(
        iter = seq_along(loglik_vec),
        loglik = loglik_vec,
        chain = factor(ch)
      )
    )
  }
}

if (nrow(loglik_df) > 0) {
  p <- ggplot(loglik_df, aes(x = iter, y = loglik, color = chain)) +
    geom_line(alpha = 0.7) +
    theme_bw() +
    labs(
      title = "Log-likelihood Trace by Chain",
      subtitle = "Block Updating with Separated Measurement Model",
      x = "Iteration",
      y = "Log-likelihood",
      color = "Chain"
    )
  ggsave("/Users/yanfei/Desktop/loglik_trace_block.png", p, width = 10, height = 6)
}

## ----Performance Metrics----
cat("\n=== PERFORMANCE SUMMARY ===\n")
cat("Model: Block Updating with Separated Measurement Model\n")
cat("Number of chains:", n_chains, "\n")
cat("Burn-in iterations:", n_burnin, "\n")
cat("Post-burnin iterations:", n_post, "\n")
cat("Total iterations per chain:", n_burnin + n_post, "\n")
cat("Total iterations across chains:", n_chains * (n_burnin + n_post), "\n")

# Calculate effective samples per second if run time is available
if (exists("results") && "timetaken" %in% names(results)) {
  total_time <- as.numeric(results$timetaken)
  cat("Total run time:", round(total_time, 1), "seconds\n")
  
  if (exists("ess")) {
    total_ess <- sum(ess, na.rm = TRUE)
    ess_per_sec <- total_ess / total_time
    cat("Total effective samples:", round(total_ess, 0), "\n")
    cat("Effective samples per second:", round(ess_per_sec, 2), "\n")
    
    # Compare with hypothetical univariate updating
    cat("\nEstimated efficiency gain:\n")
    cat("• Block updating reduces autocorrelation\n")
    cat("• Separated measurement model simplifies computation\n")
    cat("• Hierarchical shrinkage improves parameter estimation\n")
  }
}

## ----Convergence Assessment Summary----
cat("\n=== CONVERGENCE ASSESSMENT ===\n")

# Calculate R-hat for all monitored parameters
if (exists("gelman")) {
  rhat_values <- gelman$psrf[, "Point est."]
  n_converged <- sum(rhat_values < 1.1, na.rm = TRUE)
  n_total <- length(rhat_values)
  
  cat("Parameters with R-hat < 1.1:", n_converged, "/", n_total, 
      paste0("(", round(100 * n_converged/n_total, 1), "%)\n"))
  
  if (n_converged == n_total) {
    cat("✓ All parameters converged!\n")
  } else {
    cat("⚠ Some parameters need more iterations\n")
  }
}

## ----Save MCMC samples for future use----
# saveRDS(samples_full, "/Users/yanfei/Desktop/mcmc_samples_block.rds")
# saveRDS(results, "/Users/yanfei/Desktop/jags_results_block.rds")

cat("\n=== ANALYSIS COMPLETE ===\n")
cat("Results saved to:\n")
cat("1. /Users/yanfei/Desktop/mcmc_samples_block.rds\n")
cat("2. /Users/yanfei/Desktop/jags_results_block.rds\n")
cat("3. Various diagnostic plots (.png)\n")
cat("4. Parameter summaries (.csv)\n")




```

# WAIC range
```{r}






# Function to compute WAIC from posterior samples
compute_waic <- function(post_samples, data_jags) {
  # Extract lambda samples
  lambda_samples <- post_samples[, grep("^lambda", colnames(post_samples))]
  
  # Number of observations
  n_obs <- data_jags$N1
  
  # Compute log-likelihood for each posterior sample and observation
  log_lik <- matrix(NA, nrow = nrow(lambda_samples), ncol = n_obs)
  
  for (i in 1:n_obs) {
    # Poisson log-likelihood
    log_lik[, i] <- dpois(data_jags$Y_out[i], 
                          lambda = lambda_samples[, i], 
                          log = TRUE)
  }
  
  # Compute WAIC using loo package
  waic_result <- waic(log_lik)
  
  return(list(
    waic = waic_result$estimates["waic", "Estimate"],
    se = waic_result$estimates["waic", "SE"],
    p_waic = waic_result$estimates["p_waic", "Estimate"],
    log_lik = log_lik
  ))
}

# Method 1: Bootstrap WAIC interval
compute_waic_bootstrap <- function(results, data_jags, n_bootstrap = 200) {
  post <- as.matrix(as.mcmc(results))
  n_samples <- nrow(post)
  
  waic_boot <- numeric(n_bootstrap)
  
  for (b in 1:n_bootstrap) {
    # Bootstrap resample with replacement
    boot_indices <- sample(1:n_samples, size = n_samples, replace = TRUE)
    post_boot <- post[boot_indices, ]
    
    # Compute WAIC on bootstrap sample
    waic_result <- compute_waic(post_boot, data_jags)
    waic_boot[b] <- waic_result$waic
  }
  
  # Bootstrap confidence interval
  waic_ci <- quantile(waic_boot, c(0.025, 0.975))
  
  return(list(
    waic_mean = mean(waic_boot),
    waic_sd = sd(waic_boot),
    waic_ci_lower = waic_ci[1],
    waic_ci_upper = waic_ci[2],
    waic_boot = waic_boot
  ))
}





post_all <- as.matrix(as.mcmc(results))

# Compute WAIC with different methods
waic_standard <- compute_waic(post_all, data_jags)
waic_bootstrap <- compute_waic_bootstrap(results, data_jags, n_bootstrap = 100)

cat("Bootstrap WAIC (95% CI):\n")
cat("  Mean:", round(waic_bootstrap$waic_mean, 2), "\n")
cat("  SD:", round(waic_bootstrap$waic_sd, 2), "\n")
cat("  95% CI: [", round(waic_bootstrap$waic_ci_lower, 2), ", ", 
    round(waic_bootstrap$waic_ci_upper, 2), "]\n\n")

# Visualize bootstrap distribution
if (require(ggplot2)) {
  boot_df <- data.frame(waic = waic_bootstrap$waic_boot)
  
  p <- ggplot(boot_df, aes(x = waic)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
    geom_vline(xintercept = waic_bootstrap$waic_ci_lower, 
               linetype = "dashed", color = "red", size = 1) +
    geom_vline(xintercept = waic_bootstrap$waic_ci_upper, 
               linetype = "dashed", color = "red", size = 1) +
    geom_vline(xintercept = waic_standard$waic, 
               linetype = "solid", color = "darkgreen", size = 1) +
    labs(title = "Bootstrap Distribution of WAIC",
         x = "WAIC", y = "Frequency") +
    theme_minimal()
  
  print(p)
}



```
#  COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE TARGET POLLUTANT
## Extract the key info
```{r}

post <- as.matrix(as.mcmc(results))
n_samples <- nrow(post)
n_timepoints <- data_jags$N1  

# ----------------------------------------------------------------------------
# Step 1: Extract all necessary posterior samples
# ----------------------------------------------------------------------------

# Extract beta_out coefficients (pollutant effects)
beta_out_cols <- grep("^beta_out\\[", colnames(post))
beta_out_samples <- post[, beta_out_cols]  # n_samples × K matrix

# Extract gamma (temperature effect)
gamma_samples <- post[, "gamma"]

# Extract alpha (intercept)
alpha_samples <- post[, "alpha"]

# Extract muY_orig (back-transformed latent pollutants)
muY_orig_cols <- grep("^muY_orig\\[", colnames(post))
# muY_orig is stored as muY_orig[t, j] - we need to reshape
n_muY <- length(muY_orig_cols)  # Should be N × K = 258 × 6 = 1548

# Reshape muY_orig into array: [sample, time, pollutant]
muY_orig_array <- array(NA, dim = c(n_samples, n_timepoints, K))
for (s in 1:n_samples) {
  for (t in 1:n_timepoints) {
    for (j in 1:K) {
      col_name <- paste0("muY_orig[", t, ",", j, "]")
      if (col_name %in% colnames(post)) {
        muY_orig_array[s, t, j] <- post[s, col_name]
      }
    }
  }
}

# Extract offset (population)
offset <- data_jags$offset

# Extract temperature (yv)
temperature <- data_jags$yv
```

## 1st pollutant
```{r}
# Identify the target pollutant
pm10_idx <- 1

# Initialize arrays for predictions
lambda_baseline <- matrix(NA, nrow = n_samples, ncol = n_timepoints)
lambda_counter <- matrix(NA, nrow = n_samples, ncol = n_timepoints)

muY_modified <- muY_orig_array  # Copy entire array
muY_modified[, , pm10_idx] <- muY_orig_array[, , pm10_idx] + 0.1  # Add 10% on log scale

for (s in 1:n_samples) {
  
  # Extract parameters for this sample
  beta_out_s <- beta_out_samples[s, ]
  gamma_s <- gamma_samples[s]
  alpha_s <- alpha_samples[s]
  
  # For each timepoint
  for (t in 2:n_timepoints) {
    
    # ----- Scenario A: Baseline (using latent true PM10) -----
    linear_pred_base <- alpha_s +
      sum(beta_out_s * muY_orig_array[s, t-1, ]) +
      gamma_s * temperature[t]
    
    # Add offset and exponentiate
    lambda_baseline[s, t] <- exp(linear_pred_base + log(offset[t]))
    
    # ----- Scenario B: 10% increase in the target pollutant (logged scale) -----
    
    linear_pred_counter <- alpha_s +
      sum(beta_out_s * muY_modified[s, t-1, ]) +
      gamma_s * temperature[t]
    
    lambda_counter[s, t] <- exp(linear_pred_counter + log(offset[t]))
  }
  
  # Progress indicator
  if (s %% 500 == 0) cat("Processed", s, "samples\n")
}

# Absolute difference (excess deaths)
delta_absolute <- lambda_counter - lambda_baseline  # n_samples × n_timepoints

# Relative difference (percent change)
delta_relative <- (delta_absolute / lambda_baseline) * 100  # percent

# Initialize matrices for summaries
n_timepoints <- ncol(delta_absolute)
timepoint_summary <- matrix(NA, nrow = n_timepoints-1, ncol = 8)
colnames(timepoint_summary) <- c("time",
                                 "mean_absolute", "lower_absolute", "upper_absolute",
                                 "mean_relative", "lower_relative", "upper_relative",
                                 "prob_positive")

delta_absolute=delta_absolute[,-1]
delta_relative=delta_relative[,-1]

for (t in 1:n_timepoints-1) {
  
  # Absolute effect summaries
  abs_t <- delta_absolute[, t]
  mean_abs <- mean(abs_t)
  lower_abs <- quantile(abs_t, probs = 0.025)
  upper_abs <- quantile(abs_t, probs = 0.975)
  
  # Relative effect summaries
  rel_t <- delta_relative[, t]
  mean_rel <- mean(rel_t)
  lower_rel <- quantile(rel_t, probs = 0.025)
  upper_rel <- quantile(rel_t, probs = 0.975)
  
  # Probability of positive effect (absolute)
  prob_pos <- mean(abs_t > 0)
  
  timepoint_summary[t, ] <- c(t, mean_abs, lower_abs, upper_abs,
                              mean_rel, lower_rel, upper_rel, prob_pos)
}

timepoint_summary_df <- as.data.frame(timepoint_summary)


# ----------------------------------------------------------------------------
# Step 5: Aggregate summaries
# ----------------------------------------------------------------------------

# Total excess deaths over study period (per sample)
total_excess_samples <- rowSums(delta_absolute)

# Average daily excess deaths (per sample)
avg_daily_excess_samples <- rowMeans(delta_absolute)

# Average percent change (per sample) - average across timepoints
avg_percent_change_samples <- rowMeans(delta_relative)

# Summarize across samples
aggregate_summary <- data.frame(
  Metric = c(
    "Total excess deaths (258 days)",
    "Average daily excess deaths",
    "Average percent change in mortality",
    "Median percent change",
    "Maximum daily excess deaths (mean)",
    "Days with positive effect (prob > 0.95)"
  ),
  Estimate = c(
    mean(total_excess_samples),
    mean(avg_daily_excess_samples),
    mean(avg_percent_change_samples),
    median(avg_percent_change_samples),
    max(timepoint_summary_df$mean_absolute),
    sum(timepoint_summary_df$prob_positive > 0.95)
  ),
  Lower_95 = c(
    quantile(total_excess_samples, 0.025),
    quantile(avg_daily_excess_samples, 0.025),
    quantile(avg_percent_change_samples, 0.025),
    NA,
    NA,
    NA
  ),
  Upper_95 = c(
    quantile(total_excess_samples, 0.975),
    quantile(avg_daily_excess_samples, 0.975),
    quantile(avg_percent_change_samples, 0.975),
    NA,
    NA,
    NA
  )
)

# ----------------------------------------------------------------------------
# Step 7: Effect by pollution level
# ----------------------------------------------------------------------------

# Get baseline PM10 levels (latent true values)
baseline_pm10 <- matrix(NA, nrow = n_samples, ncol = n_timepoints-1)
for (s in 1:n_samples) {
  for (t in 1:n_timepoints-1) {
    baseline_pm10[s, t] <- muY_orig_array[s, t, pm10_idx]
  }
}

# Calculate mean PM10 across samples for each timepoint
mean_pm10 <- colMeans(baseline_pm10)

# Categorize into quartiles
pm10_quartiles <- cut(mean_pm10,
                      breaks = quantile(mean_pm10, probs = c(0, 0.25, 0.5, 0.75, 1)),
                      labels = c("Low", "Medium-Low", "Medium-High", "High"),
                      include.lowest = TRUE)

pm10_stratified <- data.frame()

for (level in levels(pm10_quartiles)) {
  level_idx <- which(pm10_quartiles == level)
  
  # Extract effects for this pollution level
  level_abs <- delta_absolute[, level_idx, drop = FALSE]
  level_rel <- delta_relative[, level_idx, drop = FALSE]
  
  # Average across timepoints within level for each sample
  level_avg_abs_samples <- rowMeans(level_abs)
  level_avg_rel_samples <- rowMeans(level_rel)
  
  pm10_stratified <- rbind(pm10_stratified, data.frame(
    PM10_Level = level,
    Mean_Absolute = mean(level_avg_abs_samples),
    Lower_Absolute = quantile(level_avg_abs_samples, 0.025),
    Upper_Absolute = quantile(level_avg_abs_samples, 0.975),
    Mean_Relative = mean(level_avg_rel_samples),
    Lower_Relative = quantile(level_avg_rel_samples, 0.025),
    Upper_Relative = quantile(level_avg_rel_samples, 0.975)
  ))
}

# ----------------------------------------------------------------------------
# Step 8: Visualization
# ----------------------------------------------------------------------------

# Plot 1: Time series of excess deaths
p1 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_absolute, ymax = upper_absolute),
              fill = "red", alpha = 0.3) +
  geom_line(aes(y = mean_absolute), color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Excess Mortality from 10% Increase in True NO2",
    subtitle = "Daily estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Excess deaths per day"
  )
ggsave("/Users/yanfei/Desktop/excess_mortality_timeseries.png", p1, width = 12, height = 5)

# Plot 2: Percent change time series
p2 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_relative, ymax = upper_relative),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = mean_relative), color = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Percent Change in Mortality from 10.5% Increase in True NO2",
    subtitle = "Weekly estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Percent change (%)"
  )
ggsave("/Users/yanfei/Desktop/percent_change_timeseries.png", p2, width = 12, height = 5)

# Plot 4: Effect by pollution level
p4 <- ggplot(pm10_stratified, aes(x = PM10_Level, y = Mean_Relative, fill = PM10_Level)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Lower_Relative, ymax = Upper_Relative), width = 0.2) +
  theme_bw() +
  labs(
    title = "Mortality Effect by PM10 Level",
    subtitle = "Average percent change by pollution quartile with 95% CI",
    x = "PM10 Quartile",
    y = "Percent change (%)"
  ) +
  theme(legend.position = "none")
ggsave("/Users/yanfei/Desktop/effect_by_pollution.png", p4, width = 8, height = 5)

# Plot 5: Distribution of total excess deaths
p5 <- ggplot(data.frame(total = total_excess_samples), aes(x = total)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean(total_excess_samples), color = "red", size = 1) +
  geom_vline(xintercept = quantile(total_excess_samples, c(0.025, 0.975)),
             linetype = "dashed", color = "red", size = 0.8) +
  theme_bw() +
  labs(
    title = "Posterior Distribution: Total Excess Deaths",
    subtitle = paste0("Mean: ", round(mean(total_excess_samples), 1),
                      " (95% CI: ", round(quantile(total_excess_samples, 0.025), 1),
                      " - ", round(quantile(total_excess_samples, 0.975), 1), ")"),
    x = "Total excess deaths (258 days)",
    y = "Frequency"
  )
ggsave("/Users/yanfei/Desktop/total_excess_distribution.png", p5, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Step 9: Print formatted results for report
# ----------------------------------------------------------------------------

cat("\n", paste(rep("=", 80), collapse = ""), "\n")
cat("COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE PM10\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("OVERALL EFFECTS:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
cat(sprintf("Average percent change in mortality: %.2f%% (95%% CI: %.2f%%, %.2f%%)\n",
            aggregate_summary$Estimate[3],
            aggregate_summary$Lower_95[3],
            aggregate_summary$Upper_95[3]))
cat(sprintf("Average daily excess deaths: %.2f (95%% CI: %.2f, %.2f)\n",
            aggregate_summary$Estimate[2],
            aggregate_summary$Lower_95[2],
            aggregate_summary$Upper_95[2]))
cat(sprintf("Total excess deaths over study period: %.0f (95%% CI: %.0f, %.0f)\n",
            aggregate_summary$Estimate[1],
            aggregate_summary$Lower_95[1],
            aggregate_summary$Upper_95[1]))
cat(sprintf("Days with >95%% probability of positive effect: %.0f out of %d (%.1f%%)\n",
            aggregate_summary$Estimate[6], n_timepoints,
            100 * aggregate_summary$Estimate[6] / n_timepoints))

cat("\n\nEFFECT BY PM10 LEVEL (Percent change):\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
print(pm10_stratified[, c("PM10_Level", "Mean_Relative", "Lower_Relative", "Upper_Relative")])

cat("\n\n", paste(rep("=", 80), collapse = ""), "\n")

# ----------------------------------------------------------------------------
# Step 10: Save results for reporting
# ----------------------------------------------------------------------------

# Save all results to RDS for later use
counterfactual_results <- list(
  baseline_lambda = lambda_baseline,
  counterfactual_lambda = lambda_counter,
  delta_absolute = delta_absolute,
  delta_relative = delta_relative,
  timepoint_summary = timepoint_summary_df,
  aggregate_summary = aggregate_summary,
  # seasonal_effects = seasonal_effects,
  pm10_stratified = pm10_stratified,
  total_excess_samples = total_excess_samples,
  avg_daily_excess_samples = avg_daily_excess_samples,
  avg_percent_change_samples = avg_percent_change_samples
)

saveRDS(counterfactual_results, "/Users/yanfei/Desktop/counterfactual_results.rds")

# Also save CSV tables for easy inclusion in reports
write.csv(timepoint_summary_df, "/Users/yanfei/Desktop/daily_effects.csv", row.names = FALSE)
write.csv(aggregate_summary, "/Users/yanfei/Desktop/aggregate_effects.csv", row.names = FALSE)
# write.csv(seasonal_effects, "/Users/yanfei/Desktop/seasonal_effects.csv", row.names = FALSE)
write.csv(pm10_stratified, "/Users/yanfei/Desktop/pm10_stratified_effects.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- /Users/yanfei/Desktop/counterfactual_results.rds\n")
cat("- /Users/yanfei/Desktop/daily_effects.csv\n")
cat("- /Users/yanfei/Desktop/aggregate_effects.csv\n")
# cat("- /Users/yanfei/Desktop/seasonal_effects.csv\n")
cat("- /Users/yanfei/Desktop/pm10_stratified_effects.csv\n")
cat("- Various PNG plots in /Users/yanfei/Desktop/\n")
```

## 2nd pollutant
```{r}
# Identify the target pollutant
pm10_idx <- 2

# Initialize arrays for predictions
lambda_baseline <- matrix(NA, nrow = n_samples, ncol = n_timepoints)
lambda_counter <- matrix(NA, nrow = n_samples, ncol = n_timepoints)

muY_modified <- muY_orig_array  # Copy entire array
muY_modified[, , pm10_idx] <- muY_orig_array[, , pm10_idx] + 0.1  # Add 10% on log scale

for (s in 1:n_samples) {
  
  # Extract parameters for this sample
  beta_out_s <- beta_out_samples[s, ]
  gamma_s <- gamma_samples[s]
  alpha_s <- alpha_samples[s]
  
  # For each timepoint
  for (t in 2:n_timepoints) {
    
    # ----- Scenario A: Baseline (using latent true PM10) -----
    linear_pred_base <- alpha_s +
      sum(beta_out_s * muY_orig_array[s, t-1, ]) +
      gamma_s * temperature[t]
    
    # Add offset and exponentiate
    lambda_baseline[s, t] <- exp(linear_pred_base + log(offset[t]))
    
    # ----- Scenario B: 10% increase in the target pollutant (logged scale) -----
    
    linear_pred_counter <- alpha_s +
      sum(beta_out_s * muY_modified[s, t-1, ]) +
      gamma_s * temperature[t]
    
    lambda_counter[s, t] <- exp(linear_pred_counter + log(offset[t]))
  }
  
  # Progress indicator
  if (s %% 500 == 0) cat("Processed", s, "samples\n")
}

# Absolute difference (excess deaths)
delta_absolute <- lambda_counter - lambda_baseline  # n_samples × n_timepoints

# Relative difference (percent change)
delta_relative <- (delta_absolute / lambda_baseline) * 100  # percent

# Initialize matrices for summaries
n_timepoints <- ncol(delta_absolute)
timepoint_summary <- matrix(NA, nrow = n_timepoints-1, ncol = 8)
colnames(timepoint_summary) <- c("time",
                                 "mean_absolute", "lower_absolute", "upper_absolute",
                                 "mean_relative", "lower_relative", "upper_relative",
                                 "prob_positive")

delta_absolute=delta_absolute[,-1]
delta_relative=delta_relative[,-1]

for (t in 1:n_timepoints-1) {
  
  # Absolute effect summaries
  abs_t <- delta_absolute[, t]
  mean_abs <- mean(abs_t)
  lower_abs <- quantile(abs_t, probs = 0.025)
  upper_abs <- quantile(abs_t, probs = 0.975)
  
  # Relative effect summaries
  rel_t <- delta_relative[, t]
  mean_rel <- mean(rel_t)
  lower_rel <- quantile(rel_t, probs = 0.025)
  upper_rel <- quantile(rel_t, probs = 0.975)
  
  # Probability of positive effect (absolute)
  prob_pos <- mean(abs_t > 0)
  
  timepoint_summary[t, ] <- c(t, mean_abs, lower_abs, upper_abs,
                              mean_rel, lower_rel, upper_rel, prob_pos)
}

timepoint_summary_df <- as.data.frame(timepoint_summary)


# ----------------------------------------------------------------------------
# Step 5: Aggregate summaries
# ----------------------------------------------------------------------------

# Total excess deaths over study period (per sample)
total_excess_samples <- rowSums(delta_absolute)

# Average daily excess deaths (per sample)
avg_daily_excess_samples <- rowMeans(delta_absolute)

# Average percent change (per sample) - average across timepoints
avg_percent_change_samples <- rowMeans(delta_relative)

# Summarize across samples
aggregate_summary <- data.frame(
  Metric = c(
    "Total excess deaths (258 days)",
    "Average daily excess deaths",
    "Average percent change in mortality",
    "Median percent change",
    "Maximum daily excess deaths (mean)",
    "Days with positive effect (prob > 0.95)"
  ),
  Estimate = c(
    mean(total_excess_samples),
    mean(avg_daily_excess_samples),
    mean(avg_percent_change_samples),
    median(avg_percent_change_samples),
    max(timepoint_summary_df$mean_absolute),
    sum(timepoint_summary_df$prob_positive > 0.95)
  ),
  Lower_95 = c(
    quantile(total_excess_samples, 0.025),
    quantile(avg_daily_excess_samples, 0.025),
    quantile(avg_percent_change_samples, 0.025),
    NA,
    NA,
    NA
  ),
  Upper_95 = c(
    quantile(total_excess_samples, 0.975),
    quantile(avg_daily_excess_samples, 0.975),
    quantile(avg_percent_change_samples, 0.975),
    NA,
    NA,
    NA
  )
)

# ----------------------------------------------------------------------------
# Step 7: Effect by pollution level
# ----------------------------------------------------------------------------

# Get baseline PM10 levels (latent true values)
baseline_pm10 <- matrix(NA, nrow = n_samples, ncol = n_timepoints-1)
for (s in 1:n_samples) {
  for (t in 1:n_timepoints-1) {
    baseline_pm10[s, t] <- muY_orig_array[s, t, pm10_idx]
  }
}

# Calculate mean PM10 across samples for each timepoint
mean_pm10 <- colMeans(baseline_pm10)

# Categorize into quartiles
pm10_quartiles <- cut(mean_pm10,
                      breaks = quantile(mean_pm10, probs = c(0, 0.25, 0.5, 0.75, 1)),
                      labels = c("Low", "Medium-Low", "Medium-High", "High"),
                      include.lowest = TRUE)

pm10_stratified <- data.frame()

for (level in levels(pm10_quartiles)) {
  level_idx <- which(pm10_quartiles == level)
  
  # Extract effects for this pollution level
  level_abs <- delta_absolute[, level_idx, drop = FALSE]
  level_rel <- delta_relative[, level_idx, drop = FALSE]
  
  # Average across timepoints within level for each sample
  level_avg_abs_samples <- rowMeans(level_abs)
  level_avg_rel_samples <- rowMeans(level_rel)
  
  pm10_stratified <- rbind(pm10_stratified, data.frame(
    PM10_Level = level,
    Mean_Absolute = mean(level_avg_abs_samples),
    Lower_Absolute = quantile(level_avg_abs_samples, 0.025),
    Upper_Absolute = quantile(level_avg_abs_samples, 0.975),
    Mean_Relative = mean(level_avg_rel_samples),
    Lower_Relative = quantile(level_avg_rel_samples, 0.025),
    Upper_Relative = quantile(level_avg_rel_samples, 0.975)
  ))
}

# ----------------------------------------------------------------------------
# Step 8: Visualization
# ----------------------------------------------------------------------------

# Plot 1: Time series of excess deaths
p1 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_absolute, ymax = upper_absolute),
              fill = "red", alpha = 0.3) +
  geom_line(aes(y = mean_absolute), color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Excess Mortality from 10% Increase in True CO",
    subtitle = "Daily estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Excess deaths per day"
  )
ggsave("/Users/yanfei/Desktop/excess_mortality_timeseries.png", p1, width = 12, height = 5)

# Plot 2: Percent change time series
p2 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_relative, ymax = upper_relative),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = mean_relative), color = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Percent Change in Mortality from 10.5% Increase in True CO",
    subtitle = "Weekly estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Percent change (%)"
  )
ggsave("/Users/yanfei/Desktop/percent_change_timeseries.png", p2, width = 12, height = 5)

# Plot 4: Effect by pollution level
p4 <- ggplot(pm10_stratified, aes(x = PM10_Level, y = Mean_Relative, fill = PM10_Level)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Lower_Relative, ymax = Upper_Relative), width = 0.2) +
  theme_bw() +
  labs(
    title = "Mortality Effect by PM10 Level",
    subtitle = "Average percent change by pollution quartile with 95% CI",
    x = "PM10 Quartile",
    y = "Percent change (%)"
  ) +
  theme(legend.position = "none")
ggsave("/Users/yanfei/Desktop/effect_by_pollution.png", p4, width = 8, height = 5)

# Plot 5: Distribution of total excess deaths
p5 <- ggplot(data.frame(total = total_excess_samples), aes(x = total)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean(total_excess_samples), color = "red", size = 1) +
  geom_vline(xintercept = quantile(total_excess_samples, c(0.025, 0.975)),
             linetype = "dashed", color = "red", size = 0.8) +
  theme_bw() +
  labs(
    title = "Posterior Distribution: Total Excess Deaths",
    subtitle = paste0("Mean: ", round(mean(total_excess_samples), 1),
                      " (95% CI: ", round(quantile(total_excess_samples, 0.025), 1),
                      " - ", round(quantile(total_excess_samples, 0.975), 1), ")"),
    x = "Total excess deaths (258 days)",
    y = "Frequency"
  )
ggsave("/Users/yanfei/Desktop/total_excess_distribution.png", p5, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Step 9: Print formatted results for report
# ----------------------------------------------------------------------------

cat("\n", paste(rep("=", 80), collapse = ""), "\n")
cat("COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE PM10\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("OVERALL EFFECTS:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
cat(sprintf("Average percent change in mortality: %.2f%% (95%% CI: %.2f%%, %.2f%%)\n",
            aggregate_summary$Estimate[3],
            aggregate_summary$Lower_95[3],
            aggregate_summary$Upper_95[3]))
cat(sprintf("Average daily excess deaths: %.2f (95%% CI: %.2f, %.2f)\n",
            aggregate_summary$Estimate[2],
            aggregate_summary$Lower_95[2],
            aggregate_summary$Upper_95[2]))
cat(sprintf("Total excess deaths over study period: %.0f (95%% CI: %.0f, %.0f)\n",
            aggregate_summary$Estimate[1],
            aggregate_summary$Lower_95[1],
            aggregate_summary$Upper_95[1]))
cat(sprintf("Days with >95%% probability of positive effect: %.0f out of %d (%.1f%%)\n",
            aggregate_summary$Estimate[6], n_timepoints,
            100 * aggregate_summary$Estimate[6] / n_timepoints))

cat("\n\nEFFECT BY PM10 LEVEL (Percent change):\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
print(pm10_stratified[, c("PM10_Level", "Mean_Relative", "Lower_Relative", "Upper_Relative")])

cat("\n\n", paste(rep("=", 80), collapse = ""), "\n")

# ----------------------------------------------------------------------------
# Step 10: Save results for reporting
# ----------------------------------------------------------------------------

# Save all results to RDS for later use
counterfactual_results <- list(
  baseline_lambda = lambda_baseline,
  counterfactual_lambda = lambda_counter,
  delta_absolute = delta_absolute,
  delta_relative = delta_relative,
  timepoint_summary = timepoint_summary_df,
  aggregate_summary = aggregate_summary,
  # seasonal_effects = seasonal_effects,
  pm10_stratified = pm10_stratified,
  total_excess_samples = total_excess_samples,
  avg_daily_excess_samples = avg_daily_excess_samples,
  avg_percent_change_samples = avg_percent_change_samples
)

saveRDS(counterfactual_results, "/Users/yanfei/Desktop/counterfactual_results.rds")

# Also save CSV tables for easy inclusion in reports
write.csv(timepoint_summary_df, "/Users/yanfei/Desktop/daily_effects.csv", row.names = FALSE)
write.csv(aggregate_summary, "/Users/yanfei/Desktop/aggregate_effects.csv", row.names = FALSE)
# write.csv(seasonal_effects, "/Users/yanfei/Desktop/seasonal_effects.csv", row.names = FALSE)
write.csv(pm10_stratified, "/Users/yanfei/Desktop/pm10_stratified_effects.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- /Users/yanfei/Desktop/counterfactual_results.rds\n")
cat("- /Users/yanfei/Desktop/daily_effects.csv\n")
cat("- /Users/yanfei/Desktop/aggregate_effects.csv\n")
# cat("- /Users/yanfei/Desktop/seasonal_effects.csv\n")
cat("- /Users/yanfei/Desktop/pm10_stratified_effects.csv\n")
cat("- Various PNG plots in /Users/yanfei/Desktop/\n")
```


## 3rd pollutant
```{r}
# Identify the target pollutant
pm10_idx <- 3

# Initialize arrays for predictions
lambda_baseline <- matrix(NA, nrow = n_samples, ncol = n_timepoints)
lambda_counter <- matrix(NA, nrow = n_samples, ncol = n_timepoints)

muY_modified <- muY_orig_array  # Copy entire array
muY_modified[, , pm10_idx] <- muY_orig_array[, , pm10_idx] + 0.1  # Add 10% on log scale

for (s in 1:n_samples) {
  
  # Extract parameters for this sample
  beta_out_s <- beta_out_samples[s, ]
  gamma_s <- gamma_samples[s]
  alpha_s <- alpha_samples[s]
  
  # For each timepoint
  for (t in 2:n_timepoints) {
    
    # ----- Scenario A: Baseline (using latent true PM10) -----
    linear_pred_base <- alpha_s +
      sum(beta_out_s * muY_orig_array[s, t-1, ]) +
      gamma_s * temperature[t]
    
    # Add offset and exponentiate
    lambda_baseline[s, t] <- exp(linear_pred_base + log(offset[t]))
    
    # ----- Scenario B: 10% increase in the target pollutant (logged scale) -----
    
    linear_pred_counter <- alpha_s +
      sum(beta_out_s * muY_modified[s, t-1, ]) +
      gamma_s * temperature[t]
    
    lambda_counter[s, t] <- exp(linear_pred_counter + log(offset[t]))
  }
  
  # Progress indicator
  if (s %% 500 == 0) cat("Processed", s, "samples\n")
}

# Absolute difference (excess deaths)
delta_absolute <- lambda_counter - lambda_baseline  # n_samples × n_timepoints

# Relative difference (percent change)
delta_relative <- (delta_absolute / lambda_baseline) * 100  # percent

# Initialize matrices for summaries
n_timepoints <- ncol(delta_absolute)
timepoint_summary <- matrix(NA, nrow = n_timepoints-1, ncol = 8)
colnames(timepoint_summary) <- c("time",
                                 "mean_absolute", "lower_absolute", "upper_absolute",
                                 "mean_relative", "lower_relative", "upper_relative",
                                 "prob_positive")

delta_absolute=delta_absolute[,-1]
delta_relative=delta_relative[,-1]

for (t in 1:n_timepoints-1) {
  
  # Absolute effect summaries
  abs_t <- delta_absolute[, t]
  mean_abs <- mean(abs_t)
  lower_abs <- quantile(abs_t, probs = 0.025)
  upper_abs <- quantile(abs_t, probs = 0.975)
  
  # Relative effect summaries
  rel_t <- delta_relative[, t]
  mean_rel <- mean(rel_t)
  lower_rel <- quantile(rel_t, probs = 0.025)
  upper_rel <- quantile(rel_t, probs = 0.975)
  
  # Probability of positive effect (absolute)
  prob_pos <- mean(abs_t > 0)
  
  timepoint_summary[t, ] <- c(t, mean_abs, lower_abs, upper_abs,
                              mean_rel, lower_rel, upper_rel, prob_pos)
}

timepoint_summary_df <- as.data.frame(timepoint_summary)


# ----------------------------------------------------------------------------
# Step 5: Aggregate summaries
# ----------------------------------------------------------------------------

# Total excess deaths over study period (per sample)
total_excess_samples <- rowSums(delta_absolute)

# Average daily excess deaths (per sample)
avg_daily_excess_samples <- rowMeans(delta_absolute)

# Average percent change (per sample) - average across timepoints
avg_percent_change_samples <- rowMeans(delta_relative)

# Summarize across samples
aggregate_summary <- data.frame(
  Metric = c(
    "Total excess deaths (258 days)",
    "Average daily excess deaths",
    "Average percent change in mortality",
    "Median percent change",
    "Maximum daily excess deaths (mean)",
    "Days with positive effect (prob > 0.95)"
  ),
  Estimate = c(
    mean(total_excess_samples),
    mean(avg_daily_excess_samples),
    mean(avg_percent_change_samples),
    median(avg_percent_change_samples),
    max(timepoint_summary_df$mean_absolute),
    sum(timepoint_summary_df$prob_positive > 0.95)
  ),
  Lower_95 = c(
    quantile(total_excess_samples, 0.025),
    quantile(avg_daily_excess_samples, 0.025),
    quantile(avg_percent_change_samples, 0.025),
    NA,
    NA,
    NA
  ),
  Upper_95 = c(
    quantile(total_excess_samples, 0.975),
    quantile(avg_daily_excess_samples, 0.975),
    quantile(avg_percent_change_samples, 0.975),
    NA,
    NA,
    NA
  )
)

# ----------------------------------------------------------------------------
# Step 7: Effect by pollution level
# ----------------------------------------------------------------------------

# Get baseline PM10 levels (latent true values)
baseline_pm10 <- matrix(NA, nrow = n_samples, ncol = n_timepoints-1)
for (s in 1:n_samples) {
  for (t in 1:n_timepoints-1) {
    baseline_pm10[s, t] <- muY_orig_array[s, t, pm10_idx]
  }
}

# Calculate mean PM10 across samples for each timepoint
mean_pm10 <- colMeans(baseline_pm10)

# Categorize into quartiles
pm10_quartiles <- cut(mean_pm10,
                      breaks = quantile(mean_pm10, probs = c(0, 0.25, 0.5, 0.75, 1)),
                      labels = c("Low", "Medium-Low", "Medium-High", "High"),
                      include.lowest = TRUE)

pm10_stratified <- data.frame()

for (level in levels(pm10_quartiles)) {
  level_idx <- which(pm10_quartiles == level)
  
  # Extract effects for this pollution level
  level_abs <- delta_absolute[, level_idx, drop = FALSE]
  level_rel <- delta_relative[, level_idx, drop = FALSE]
  
  # Average across timepoints within level for each sample
  level_avg_abs_samples <- rowMeans(level_abs)
  level_avg_rel_samples <- rowMeans(level_rel)
  
  pm10_stratified <- rbind(pm10_stratified, data.frame(
    PM10_Level = level,
    Mean_Absolute = mean(level_avg_abs_samples),
    Lower_Absolute = quantile(level_avg_abs_samples, 0.025),
    Upper_Absolute = quantile(level_avg_abs_samples, 0.975),
    Mean_Relative = mean(level_avg_rel_samples),
    Lower_Relative = quantile(level_avg_rel_samples, 0.025),
    Upper_Relative = quantile(level_avg_rel_samples, 0.975)
  ))
}

# ----------------------------------------------------------------------------
# Step 8: Visualization
# ----------------------------------------------------------------------------

# Plot 1: Time series of excess deaths
p1 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_absolute, ymax = upper_absolute),
              fill = "red", alpha = 0.3) +
  geom_line(aes(y = mean_absolute), color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Excess Mortality from 10% Increase in True O3",
    subtitle = "Daily estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Excess deaths per day"
  )
ggsave("/Users/yanfei/Desktop/excess_mortality_timeseries.png", p1, width = 12, height = 5)

# Plot 2: Percent change time series
p2 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_relative, ymax = upper_relative),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = mean_relative), color = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Percent Change in Mortality from 10.5% Increase in True O3",
    subtitle = "Weekly estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Percent change (%)"
  )
ggsave("/Users/yanfei/Desktop/percent_change_timeseries.png", p2, width = 12, height = 5)

# Plot 4: Effect by pollution level
p4 <- ggplot(pm10_stratified, aes(x = PM10_Level, y = Mean_Relative, fill = PM10_Level)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Lower_Relative, ymax = Upper_Relative), width = 0.2) +
  theme_bw() +
  labs(
    title = "Mortality Effect by PM10 Level",
    subtitle = "Average percent change by pollution quartile with 95% CI",
    x = "PM10 Quartile",
    y = "Percent change (%)"
  ) +
  theme(legend.position = "none")
ggsave("/Users/yanfei/Desktop/effect_by_pollution.png", p4, width = 8, height = 5)

# Plot 5: Distribution of total excess deaths
p5 <- ggplot(data.frame(total = total_excess_samples), aes(x = total)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean(total_excess_samples), color = "red", size = 1) +
  geom_vline(xintercept = quantile(total_excess_samples, c(0.025, 0.975)),
             linetype = "dashed", color = "red", size = 0.8) +
  theme_bw() +
  labs(
    title = "Posterior Distribution: Total Excess Deaths",
    subtitle = paste0("Mean: ", round(mean(total_excess_samples), 1),
                      " (95% CI: ", round(quantile(total_excess_samples, 0.025), 1),
                      " - ", round(quantile(total_excess_samples, 0.975), 1), ")"),
    x = "Total excess deaths (258 days)",
    y = "Frequency"
  )
ggsave("/Users/yanfei/Desktop/total_excess_distribution.png", p5, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Step 9: Print formatted results for report
# ----------------------------------------------------------------------------

cat("\n", paste(rep("=", 80), collapse = ""), "\n")
cat("COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE PM10\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("OVERALL EFFECTS:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
cat(sprintf("Average percent change in mortality: %.2f%% (95%% CI: %.2f%%, %.2f%%)\n",
            aggregate_summary$Estimate[3],
            aggregate_summary$Lower_95[3],
            aggregate_summary$Upper_95[3]))
cat(sprintf("Average daily excess deaths: %.2f (95%% CI: %.2f, %.2f)\n",
            aggregate_summary$Estimate[2],
            aggregate_summary$Lower_95[2],
            aggregate_summary$Upper_95[2]))
cat(sprintf("Total excess deaths over study period: %.0f (95%% CI: %.0f, %.0f)\n",
            aggregate_summary$Estimate[1],
            aggregate_summary$Lower_95[1],
            aggregate_summary$Upper_95[1]))
cat(sprintf("Days with >95%% probability of positive effect: %.0f out of %d (%.1f%%)\n",
            aggregate_summary$Estimate[6], n_timepoints,
            100 * aggregate_summary$Estimate[6] / n_timepoints))

cat("\n\nEFFECT BY PM10 LEVEL (Percent change):\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
print(pm10_stratified[, c("PM10_Level", "Mean_Relative", "Lower_Relative", "Upper_Relative")])

cat("\n\n", paste(rep("=", 80), collapse = ""), "\n")

# ----------------------------------------------------------------------------
# Step 10: Save results for reporting
# ----------------------------------------------------------------------------

# Save all results to RDS for later use
counterfactual_results <- list(
  baseline_lambda = lambda_baseline,
  counterfactual_lambda = lambda_counter,
  delta_absolute = delta_absolute,
  delta_relative = delta_relative,
  timepoint_summary = timepoint_summary_df,
  aggregate_summary = aggregate_summary,
  # seasonal_effects = seasonal_effects,
  pm10_stratified = pm10_stratified,
  total_excess_samples = total_excess_samples,
  avg_daily_excess_samples = avg_daily_excess_samples,
  avg_percent_change_samples = avg_percent_change_samples
)

saveRDS(counterfactual_results, "/Users/yanfei/Desktop/counterfactual_results.rds")

# Also save CSV tables for easy inclusion in reports
write.csv(timepoint_summary_df, "/Users/yanfei/Desktop/daily_effects.csv", row.names = FALSE)
write.csv(aggregate_summary, "/Users/yanfei/Desktop/aggregate_effects.csv", row.names = FALSE)
# write.csv(seasonal_effects, "/Users/yanfei/Desktop/seasonal_effects.csv", row.names = FALSE)
write.csv(pm10_stratified, "/Users/yanfei/Desktop/pm10_stratified_effects.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- /Users/yanfei/Desktop/counterfactual_results.rds\n")
cat("- /Users/yanfei/Desktop/daily_effects.csv\n")
cat("- /Users/yanfei/Desktop/aggregate_effects.csv\n")
# cat("- /Users/yanfei/Desktop/seasonal_effects.csv\n")
cat("- /Users/yanfei/Desktop/pm10_stratified_effects.csv\n")
cat("- Various PNG plots in /Users/yanfei/Desktop/\n")
```

## 4th pollutant
```{r}
# Identify the target pollutant
pm10_idx <- 4

# Initialize arrays for predictions
lambda_baseline <- matrix(NA, nrow = n_samples, ncol = n_timepoints)
lambda_counter <- matrix(NA, nrow = n_samples, ncol = n_timepoints)

muY_modified <- muY_orig_array  # Copy entire array
muY_modified[, , pm10_idx] <- muY_orig_array[, , pm10_idx] + 0.1  # Add 10% on log scale

for (s in 1:n_samples) {
  
  # Extract parameters for this sample
  beta_out_s <- beta_out_samples[s, ]
  gamma_s <- gamma_samples[s]
  alpha_s <- alpha_samples[s]
  
  # For each timepoint
  for (t in 2:n_timepoints) {
    
    # ----- Scenario A: Baseline (using latent true PM10) -----
    linear_pred_base <- alpha_s +
      sum(beta_out_s * muY_orig_array[s, t-1, ]) +
      gamma_s * temperature[t]
    
    # Add offset and exponentiate
    lambda_baseline[s, t] <- exp(linear_pred_base + log(offset[t]))
    
    # ----- Scenario B: 10% increase in the target pollutant (logged scale) -----
    
    linear_pred_counter <- alpha_s +
      sum(beta_out_s * muY_modified[s, t-1, ]) +
      gamma_s * temperature[t]
    
    lambda_counter[s, t] <- exp(linear_pred_counter + log(offset[t]))
  }
  
  # Progress indicator
  if (s %% 500 == 0) cat("Processed", s, "samples\n")
}

# Absolute difference (excess deaths)
delta_absolute <- lambda_counter - lambda_baseline  # n_samples × n_timepoints

# Relative difference (percent change)
delta_relative <- (delta_absolute / lambda_baseline) * 100  # percent

# Initialize matrices for summaries
n_timepoints <- ncol(delta_absolute)
timepoint_summary <- matrix(NA, nrow = n_timepoints-1, ncol = 8)
colnames(timepoint_summary) <- c("time",
                                 "mean_absolute", "lower_absolute", "upper_absolute",
                                 "mean_relative", "lower_relative", "upper_relative",
                                 "prob_positive")

delta_absolute=delta_absolute[,-1]
delta_relative=delta_relative[,-1]

for (t in 1:n_timepoints-1) {
  
  # Absolute effect summaries
  abs_t <- delta_absolute[, t]
  mean_abs <- mean(abs_t)
  lower_abs <- quantile(abs_t, probs = 0.025)
  upper_abs <- quantile(abs_t, probs = 0.975)
  
  # Relative effect summaries
  rel_t <- delta_relative[, t]
  mean_rel <- mean(rel_t)
  lower_rel <- quantile(rel_t, probs = 0.025)
  upper_rel <- quantile(rel_t, probs = 0.975)
  
  # Probability of positive effect (absolute)
  prob_pos <- mean(abs_t > 0)
  
  timepoint_summary[t, ] <- c(t, mean_abs, lower_abs, upper_abs,
                              mean_rel, lower_rel, upper_rel, prob_pos)
}

timepoint_summary_df <- as.data.frame(timepoint_summary)


# ----------------------------------------------------------------------------
# Step 5: Aggregate summaries
# ----------------------------------------------------------------------------

# Total excess deaths over study period (per sample)
total_excess_samples <- rowSums(delta_absolute)

# Average daily excess deaths (per sample)
avg_daily_excess_samples <- rowMeans(delta_absolute)

# Average percent change (per sample) - average across timepoints
avg_percent_change_samples <- rowMeans(delta_relative)

# Summarize across samples
aggregate_summary <- data.frame(
  Metric = c(
    "Total excess deaths (258 days)",
    "Average daily excess deaths",
    "Average percent change in mortality",
    "Median percent change",
    "Maximum daily excess deaths (mean)",
    "Days with positive effect (prob > 0.95)"
  ),
  Estimate = c(
    mean(total_excess_samples),
    mean(avg_daily_excess_samples),
    mean(avg_percent_change_samples),
    median(avg_percent_change_samples),
    max(timepoint_summary_df$mean_absolute),
    sum(timepoint_summary_df$prob_positive > 0.95)
  ),
  Lower_95 = c(
    quantile(total_excess_samples, 0.025),
    quantile(avg_daily_excess_samples, 0.025),
    quantile(avg_percent_change_samples, 0.025),
    NA,
    NA,
    NA
  ),
  Upper_95 = c(
    quantile(total_excess_samples, 0.975),
    quantile(avg_daily_excess_samples, 0.975),
    quantile(avg_percent_change_samples, 0.975),
    NA,
    NA,
    NA
  )
)

# ----------------------------------------------------------------------------
# Step 7: Effect by pollution level
# ----------------------------------------------------------------------------

# Get baseline PM10 levels (latent true values)
baseline_pm10 <- matrix(NA, nrow = n_samples, ncol = n_timepoints-1)
for (s in 1:n_samples) {
  for (t in 1:n_timepoints-1) {
    baseline_pm10[s, t] <- muY_orig_array[s, t, pm10_idx]
  }
}

# Calculate mean PM10 across samples for each timepoint
mean_pm10 <- colMeans(baseline_pm10)

# Categorize into quartiles
pm10_quartiles <- cut(mean_pm10,
                      breaks = quantile(mean_pm10, probs = c(0, 0.25, 0.5, 0.75, 1)),
                      labels = c("Low", "Medium-Low", "Medium-High", "High"),
                      include.lowest = TRUE)

pm10_stratified <- data.frame()

for (level in levels(pm10_quartiles)) {
  level_idx <- which(pm10_quartiles == level)
  
  # Extract effects for this pollution level
  level_abs <- delta_absolute[, level_idx, drop = FALSE]
  level_rel <- delta_relative[, level_idx, drop = FALSE]
  
  # Average across timepoints within level for each sample
  level_avg_abs_samples <- rowMeans(level_abs)
  level_avg_rel_samples <- rowMeans(level_rel)
  
  pm10_stratified <- rbind(pm10_stratified, data.frame(
    PM10_Level = level,
    Mean_Absolute = mean(level_avg_abs_samples),
    Lower_Absolute = quantile(level_avg_abs_samples, 0.025),
    Upper_Absolute = quantile(level_avg_abs_samples, 0.975),
    Mean_Relative = mean(level_avg_rel_samples),
    Lower_Relative = quantile(level_avg_rel_samples, 0.025),
    Upper_Relative = quantile(level_avg_rel_samples, 0.975)
  ))
}

# ----------------------------------------------------------------------------
# Step 8: Visualization
# ----------------------------------------------------------------------------

# Plot 1: Time series of excess deaths
p1 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_absolute, ymax = upper_absolute),
              fill = "red", alpha = 0.3) +
  geom_line(aes(y = mean_absolute), color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Excess Mortality from 10% Increase in True CO",
    subtitle = "Daily estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Excess deaths per day"
  )
ggsave("/Users/yanfei/Desktop/excess_mortality_timeseries.png", p1, width = 12, height = 5)

# Plot 2: Percent change time series
p2 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_relative, ymax = upper_relative),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = mean_relative), color = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Percent Change in Mortality from 10.5% Increase in True CO",
    subtitle = "Weekly estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Percent change (%)"
  )
ggsave("/Users/yanfei/Desktop/percent_change_timeseries.png", p2, width = 12, height = 5)

# Plot 4: Effect by pollution level
p4 <- ggplot(pm10_stratified, aes(x = PM10_Level, y = Mean_Relative, fill = PM10_Level)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Lower_Relative, ymax = Upper_Relative), width = 0.2) +
  theme_bw() +
  labs(
    title = "Mortality Effect by PM10 Level",
    subtitle = "Average percent change by pollution quartile with 95% CI",
    x = "PM10 Quartile",
    y = "Percent change (%)"
  ) +
  theme(legend.position = "none")
ggsave("/Users/yanfei/Desktop/effect_by_pollution.png", p4, width = 8, height = 5)

# Plot 5: Distribution of total excess deaths
p5 <- ggplot(data.frame(total = total_excess_samples), aes(x = total)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean(total_excess_samples), color = "red", size = 1) +
  geom_vline(xintercept = quantile(total_excess_samples, c(0.025, 0.975)),
             linetype = "dashed", color = "red", size = 0.8) +
  theme_bw() +
  labs(
    title = "Posterior Distribution: Total Excess Deaths",
    subtitle = paste0("Mean: ", round(mean(total_excess_samples), 1),
                      " (95% CI: ", round(quantile(total_excess_samples, 0.025), 1),
                      " - ", round(quantile(total_excess_samples, 0.975), 1), ")"),
    x = "Total excess deaths (258 days)",
    y = "Frequency"
  )
ggsave("/Users/yanfei/Desktop/total_excess_distribution.png", p5, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Step 9: Print formatted results for report
# ----------------------------------------------------------------------------

cat("\n", paste(rep("=", 80), collapse = ""), "\n")
cat("COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE PM10\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("OVERALL EFFECTS:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
cat(sprintf("Average percent change in mortality: %.2f%% (95%% CI: %.2f%%, %.2f%%)\n",
            aggregate_summary$Estimate[3],
            aggregate_summary$Lower_95[3],
            aggregate_summary$Upper_95[3]))
cat(sprintf("Average daily excess deaths: %.2f (95%% CI: %.2f, %.2f)\n",
            aggregate_summary$Estimate[2],
            aggregate_summary$Lower_95[2],
            aggregate_summary$Upper_95[2]))
cat(sprintf("Total excess deaths over study period: %.0f (95%% CI: %.0f, %.0f)\n",
            aggregate_summary$Estimate[1],
            aggregate_summary$Lower_95[1],
            aggregate_summary$Upper_95[1]))
cat(sprintf("Days with >95%% probability of positive effect: %.0f out of %d (%.1f%%)\n",
            aggregate_summary$Estimate[6], n_timepoints,
            100 * aggregate_summary$Estimate[6] / n_timepoints))

cat("\n\nEFFECT BY PM10 LEVEL (Percent change):\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
print(pm10_stratified[, c("PM10_Level", "Mean_Relative", "Lower_Relative", "Upper_Relative")])

cat("\n\n", paste(rep("=", 80), collapse = ""), "\n")

# ----------------------------------------------------------------------------
# Step 10: Save results for reporting
# ----------------------------------------------------------------------------

# Save all results to RDS for later use
counterfactual_results <- list(
  baseline_lambda = lambda_baseline,
  counterfactual_lambda = lambda_counter,
  delta_absolute = delta_absolute,
  delta_relative = delta_relative,
  timepoint_summary = timepoint_summary_df,
  aggregate_summary = aggregate_summary,
  # seasonal_effects = seasonal_effects,
  pm10_stratified = pm10_stratified,
  total_excess_samples = total_excess_samples,
  avg_daily_excess_samples = avg_daily_excess_samples,
  avg_percent_change_samples = avg_percent_change_samples
)

saveRDS(counterfactual_results, "/Users/yanfei/Desktop/counterfactual_results.rds")

# Also save CSV tables for easy inclusion in reports
write.csv(timepoint_summary_df, "/Users/yanfei/Desktop/daily_effects.csv", row.names = FALSE)
write.csv(aggregate_summary, "/Users/yanfei/Desktop/aggregate_effects.csv", row.names = FALSE)
# write.csv(seasonal_effects, "/Users/yanfei/Desktop/seasonal_effects.csv", row.names = FALSE)
write.csv(pm10_stratified, "/Users/yanfei/Desktop/pm10_stratified_effects.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- /Users/yanfei/Desktop/counterfactual_results.rds\n")
cat("- /Users/yanfei/Desktop/daily_effects.csv\n")
cat("- /Users/yanfei/Desktop/aggregate_effects.csv\n")
# cat("- /Users/yanfei/Desktop/seasonal_effects.csv\n")
cat("- /Users/yanfei/Desktop/pm10_stratified_effects.csv\n")
cat("- Various PNG plots in /Users/yanfei/Desktop/\n")
```


## 5th pollutant
```{r}
# Identify the target pollutant
pm10_idx <- 5

# Initialize arrays for predictions
lambda_baseline <- matrix(NA, nrow = n_samples, ncol = n_timepoints)
lambda_counter <- matrix(NA, nrow = n_samples, ncol = n_timepoints)

muY_modified <- muY_orig_array  # Copy entire array
muY_modified[, , pm10_idx] <- muY_orig_array[, , pm10_idx] + 0.1  # Add 10% on log scale

for (s in 1:n_samples) {
  
  # Extract parameters for this sample
  beta_out_s <- beta_out_samples[s, ]
  gamma_s <- gamma_samples[s]
  alpha_s <- alpha_samples[s]
  
  # For each timepoint
  for (t in 2:n_timepoints) {
    
    # ----- Scenario A: Baseline (using latent true PM10) -----
    linear_pred_base <- alpha_s +
      sum(beta_out_s * muY_orig_array[s, t-1, ]) +
      gamma_s * temperature[t]
    
    # Add offset and exponentiate
    lambda_baseline[s, t] <- exp(linear_pred_base + log(offset[t]))
    
    # ----- Scenario B: 10% increase in the target pollutant (logged scale) -----
    
    linear_pred_counter <- alpha_s +
      sum(beta_out_s * muY_modified[s, t-1, ]) +
      gamma_s * temperature[t]
    
    lambda_counter[s, t] <- exp(linear_pred_counter + log(offset[t]))
  }
  
  # Progress indicator
  if (s %% 500 == 0) cat("Processed", s, "samples\n")
}

# Absolute difference (excess deaths)
delta_absolute <- lambda_counter - lambda_baseline  # n_samples × n_timepoints

# Relative difference (percent change)
delta_relative <- (delta_absolute / lambda_baseline) * 100  # percent

# Initialize matrices for summaries
n_timepoints <- ncol(delta_absolute)
timepoint_summary <- matrix(NA, nrow = n_timepoints-1, ncol = 8)
colnames(timepoint_summary) <- c("time",
                                 "mean_absolute", "lower_absolute", "upper_absolute",
                                 "mean_relative", "lower_relative", "upper_relative",
                                 "prob_positive")

delta_absolute=delta_absolute[,-1]
delta_relative=delta_relative[,-1]

for (t in 1:n_timepoints-1) {
  
  # Absolute effect summaries
  abs_t <- delta_absolute[, t]
  mean_abs <- mean(abs_t)
  lower_abs <- quantile(abs_t, probs = 0.025)
  upper_abs <- quantile(abs_t, probs = 0.975)
  
  # Relative effect summaries
  rel_t <- delta_relative[, t]
  mean_rel <- mean(rel_t)
  lower_rel <- quantile(rel_t, probs = 0.025)
  upper_rel <- quantile(rel_t, probs = 0.975)
  
  # Probability of positive effect (absolute)
  prob_pos <- mean(abs_t > 0)
  
  timepoint_summary[t, ] <- c(t, mean_abs, lower_abs, upper_abs,
                              mean_rel, lower_rel, upper_rel, prob_pos)
}

timepoint_summary_df <- as.data.frame(timepoint_summary)


# ----------------------------------------------------------------------------
# Step 5: Aggregate summaries
# ----------------------------------------------------------------------------

# Total excess deaths over study period (per sample)
total_excess_samples <- rowSums(delta_absolute)

# Average daily excess deaths (per sample)
avg_daily_excess_samples <- rowMeans(delta_absolute)

# Average percent change (per sample) - average across timepoints
avg_percent_change_samples <- rowMeans(delta_relative)

# Summarize across samples
aggregate_summary <- data.frame(
  Metric = c(
    "Total excess deaths (258 days)",
    "Average daily excess deaths",
    "Average percent change in mortality",
    "Median percent change",
    "Maximum daily excess deaths (mean)",
    "Days with positive effect (prob > 0.95)"
  ),
  Estimate = c(
    mean(total_excess_samples),
    mean(avg_daily_excess_samples),
    mean(avg_percent_change_samples),
    median(avg_percent_change_samples),
    max(timepoint_summary_df$mean_absolute),
    sum(timepoint_summary_df$prob_positive > 0.95)
  ),
  Lower_95 = c(
    quantile(total_excess_samples, 0.025),
    quantile(avg_daily_excess_samples, 0.025),
    quantile(avg_percent_change_samples, 0.025),
    NA,
    NA,
    NA
  ),
  Upper_95 = c(
    quantile(total_excess_samples, 0.975),
    quantile(avg_daily_excess_samples, 0.975),
    quantile(avg_percent_change_samples, 0.975),
    NA,
    NA,
    NA
  )
)

# ----------------------------------------------------------------------------
# Step 7: Effect by pollution level
# ----------------------------------------------------------------------------

# Get baseline PM10 levels (latent true values)
baseline_pm10 <- matrix(NA, nrow = n_samples, ncol = n_timepoints-1)
for (s in 1:n_samples) {
  for (t in 1:n_timepoints-1) {
    baseline_pm10[s, t] <- muY_orig_array[s, t, pm10_idx]
  }
}

# Calculate mean PM10 across samples for each timepoint
mean_pm10 <- colMeans(baseline_pm10)

# Categorize into quartiles
pm10_quartiles <- cut(mean_pm10,
                      breaks = quantile(mean_pm10, probs = c(0, 0.25, 0.5, 0.75, 1)),
                      labels = c("Low", "Medium-Low", "Medium-High", "High"),
                      include.lowest = TRUE)

pm10_stratified <- data.frame()

for (level in levels(pm10_quartiles)) {
  level_idx <- which(pm10_quartiles == level)
  
  # Extract effects for this pollution level
  level_abs <- delta_absolute[, level_idx, drop = FALSE]
  level_rel <- delta_relative[, level_idx, drop = FALSE]
  
  # Average across timepoints within level for each sample
  level_avg_abs_samples <- rowMeans(level_abs)
  level_avg_rel_samples <- rowMeans(level_rel)
  
  pm10_stratified <- rbind(pm10_stratified, data.frame(
    PM10_Level = level,
    Mean_Absolute = mean(level_avg_abs_samples),
    Lower_Absolute = quantile(level_avg_abs_samples, 0.025),
    Upper_Absolute = quantile(level_avg_abs_samples, 0.975),
    Mean_Relative = mean(level_avg_rel_samples),
    Lower_Relative = quantile(level_avg_rel_samples, 0.025),
    Upper_Relative = quantile(level_avg_rel_samples, 0.975)
  ))
}

# ----------------------------------------------------------------------------
# Step 8: Visualization
# ----------------------------------------------------------------------------

# Plot 1: Time series of excess deaths
p1 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_absolute, ymax = upper_absolute),
              fill = "red", alpha = 0.3) +
  geom_line(aes(y = mean_absolute), color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Excess Mortality from 10% Increase in True O3",
    subtitle = "Daily estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Excess deaths per day"
  )
ggsave("/Users/yanfei/Desktop/excess_mortality_timeseries.png", p1, width = 12, height = 5)

# Plot 2: Percent change time series
p2 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_relative, ymax = upper_relative),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = mean_relative), color = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Percent Change in Mortality from 10.5% Increase in True O3",
    subtitle = "Weekly estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Percent change (%)"
  )
ggsave("/Users/yanfei/Desktop/percent_change_timeseries.png", p2, width = 12, height = 5)

# Plot 4: Effect by pollution level
p4 <- ggplot(pm10_stratified, aes(x = PM10_Level, y = Mean_Relative, fill = PM10_Level)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Lower_Relative, ymax = Upper_Relative), width = 0.2) +
  theme_bw() +
  labs(
    title = "Mortality Effect by PM10 Level",
    subtitle = "Average percent change by pollution quartile with 95% CI",
    x = "PM10 Quartile",
    y = "Percent change (%)"
  ) +
  theme(legend.position = "none")
ggsave("/Users/yanfei/Desktop/effect_by_pollution.png", p4, width = 8, height = 5)

# Plot 5: Distribution of total excess deaths
p5 <- ggplot(data.frame(total = total_excess_samples), aes(x = total)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean(total_excess_samples), color = "red", size = 1) +
  geom_vline(xintercept = quantile(total_excess_samples, c(0.025, 0.975)),
             linetype = "dashed", color = "red", size = 0.8) +
  theme_bw() +
  labs(
    title = "Posterior Distribution: Total Excess Deaths",
    subtitle = paste0("Mean: ", round(mean(total_excess_samples), 1),
                      " (95% CI: ", round(quantile(total_excess_samples, 0.025), 1),
                      " - ", round(quantile(total_excess_samples, 0.975), 1), ")"),
    x = "Total excess deaths (258 days)",
    y = "Frequency"
  )
ggsave("/Users/yanfei/Desktop/total_excess_distribution.png", p5, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Step 9: Print formatted results for report
# ----------------------------------------------------------------------------

cat("\n", paste(rep("=", 80), collapse = ""), "\n")
cat("COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE PM10\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("OVERALL EFFECTS:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
cat(sprintf("Average percent change in mortality: %.2f%% (95%% CI: %.2f%%, %.2f%%)\n",
            aggregate_summary$Estimate[3],
            aggregate_summary$Lower_95[3],
            aggregate_summary$Upper_95[3]))
cat(sprintf("Average daily excess deaths: %.2f (95%% CI: %.2f, %.2f)\n",
            aggregate_summary$Estimate[2],
            aggregate_summary$Lower_95[2],
            aggregate_summary$Upper_95[2]))
cat(sprintf("Total excess deaths over study period: %.0f (95%% CI: %.0f, %.0f)\n",
            aggregate_summary$Estimate[1],
            aggregate_summary$Lower_95[1],
            aggregate_summary$Upper_95[1]))
cat(sprintf("Days with >95%% probability of positive effect: %.0f out of %d (%.1f%%)\n",
            aggregate_summary$Estimate[6], n_timepoints,
            100 * aggregate_summary$Estimate[6] / n_timepoints))

cat("\n\nEFFECT BY PM10 LEVEL (Percent change):\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
print(pm10_stratified[, c("PM10_Level", "Mean_Relative", "Lower_Relative", "Upper_Relative")])

cat("\n\n", paste(rep("=", 80), collapse = ""), "\n")

# ----------------------------------------------------------------------------
# Step 10: Save results for reporting
# ----------------------------------------------------------------------------

# Save all results to RDS for later use
counterfactual_results <- list(
  baseline_lambda = lambda_baseline,
  counterfactual_lambda = lambda_counter,
  delta_absolute = delta_absolute,
  delta_relative = delta_relative,
  timepoint_summary = timepoint_summary_df,
  aggregate_summary = aggregate_summary,
  # seasonal_effects = seasonal_effects,
  pm10_stratified = pm10_stratified,
  total_excess_samples = total_excess_samples,
  avg_daily_excess_samples = avg_daily_excess_samples,
  avg_percent_change_samples = avg_percent_change_samples
)

saveRDS(counterfactual_results, "/Users/yanfei/Desktop/counterfactual_results.rds")

# Also save CSV tables for easy inclusion in reports
write.csv(timepoint_summary_df, "/Users/yanfei/Desktop/daily_effects.csv", row.names = FALSE)
write.csv(aggregate_summary, "/Users/yanfei/Desktop/aggregate_effects.csv", row.names = FALSE)
# write.csv(seasonal_effects, "/Users/yanfei/Desktop/seasonal_effects.csv", row.names = FALSE)
write.csv(pm10_stratified, "/Users/yanfei/Desktop/pm10_stratified_effects.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- /Users/yanfei/Desktop/counterfactual_results.rds\n")
cat("- /Users/yanfei/Desktop/daily_effects.csv\n")
cat("- /Users/yanfei/Desktop/aggregate_effects.csv\n")
# cat("- /Users/yanfei/Desktop/seasonal_effects.csv\n")
cat("- /Users/yanfei/Desktop/pm10_stratified_effects.csv\n")
cat("- Various PNG plots in /Users/yanfei/Desktop/\n")
```
## Temperature
```{r}
# Identify the target pollutant
# pm10_idx <- 2

# Initialize arrays for predictions
lambda_baseline <- matrix(NA, nrow = n_samples, ncol = n_timepoints)
lambda_counter <- matrix(NA, nrow = n_samples, ncol = n_timepoints)

muY_modified <- muY_orig_array  # Copy entire array
# muY_modified[, , pm10_idx] <- muY_orig_array[, , pm10_idx] + 0.1  # Add 10% on log scale
temperature_modified=temperature+1
for (s in 1:n_samples) {
  
  # Extract parameters for this sample
  beta_out_s <- beta_out_samples[s, ]
  gamma_s <- gamma_samples[s]
  alpha_s <- alpha_samples[s]
  
  # For each timepoint
  for (t in 2:n_timepoints) {
    
    # ----- Scenario A: Baseline (using latent true PM10) -----
    linear_pred_base <- alpha_s +
      sum(beta_out_s * muY_orig_array[s, t-1, ]) +
      gamma_s * temperature[t]
    
    # Add offset and exponentiate
    lambda_baseline[s, t] <- exp(linear_pred_base + log(offset[t]))
    
    # ----- Scenario B: 10% increase in the target pollutant (logged scale) -----
    
    linear_pred_counter <- alpha_s +
      sum(beta_out_s * muY_modified[s, t-1, ]) +
      gamma_s * temperature_modified[t]
    
    lambda_counter[s, t] <- exp(linear_pred_counter + log(offset[t]))
  }
  
  # Progress indicator
  if (s %% 500 == 0) cat("Processed", s, "samples\n")
}

# Absolute difference (excess deaths)
delta_absolute <- lambda_counter - lambda_baseline  # n_samples × n_timepoints

# Relative difference (percent change)
delta_relative <- (delta_absolute / lambda_baseline) * 100  # percent

# Initialize matrices for summaries
n_timepoints <- ncol(delta_absolute)
timepoint_summary <- matrix(NA, nrow = n_timepoints-1, ncol = 8)
colnames(timepoint_summary) <- c("time",
                                 "mean_absolute", "lower_absolute", "upper_absolute",
                                 "mean_relative", "lower_relative", "upper_relative",
                                 "prob_positive")

delta_absolute=delta_absolute[,-1]
delta_relative=delta_relative[,-1]

for (t in 1:n_timepoints-1) {
  
  # Absolute effect summaries
  abs_t <- delta_absolute[, t]
  mean_abs <- mean(abs_t)
  lower_abs <- quantile(abs_t, probs = 0.025)
  upper_abs <- quantile(abs_t, probs = 0.975)
  
  # Relative effect summaries
  rel_t <- delta_relative[, t]
  mean_rel <- mean(rel_t)
  lower_rel <- quantile(rel_t, probs = 0.025)
  upper_rel <- quantile(rel_t, probs = 0.975)
  
  # Probability of positive effect (absolute)
  prob_pos <- mean(abs_t > 0)
  
  timepoint_summary[t, ] <- c(t, mean_abs, lower_abs, upper_abs,
                              mean_rel, lower_rel, upper_rel, prob_pos)
}

timepoint_summary_df <- as.data.frame(timepoint_summary)


# ----------------------------------------------------------------------------
# Step 5: Aggregate summaries
# ----------------------------------------------------------------------------

# Total excess deaths over study period (per sample)
total_excess_samples <- rowSums(delta_absolute)

# Average daily excess deaths (per sample)
avg_daily_excess_samples <- rowMeans(delta_absolute)

# Average percent change (per sample) - average across timepoints
avg_percent_change_samples <- rowMeans(delta_relative)

# Summarize across samples
aggregate_summary <- data.frame(
  Metric = c(
    "Total excess deaths (258 days)",
    "Average daily excess deaths",
    "Average percent change in mortality",
    "Median percent change",
    "Maximum daily excess deaths (mean)",
    "Days with positive effect (prob > 0.95)"
  ),
  Estimate = c(
    mean(total_excess_samples),
    mean(avg_daily_excess_samples),
    mean(avg_percent_change_samples),
    median(avg_percent_change_samples),
    max(timepoint_summary_df$mean_absolute),
    sum(timepoint_summary_df$prob_positive > 0.95)
  ),
  Lower_95 = c(
    quantile(total_excess_samples, 0.025),
    quantile(avg_daily_excess_samples, 0.025),
    quantile(avg_percent_change_samples, 0.025),
    NA,
    NA,
    NA
  ),
  Upper_95 = c(
    quantile(total_excess_samples, 0.975),
    quantile(avg_daily_excess_samples, 0.975),
    quantile(avg_percent_change_samples, 0.975),
    NA,
    NA,
    NA
  )
)

# ----------------------------------------------------------------------------
# Step 7: Effect by pollution level
# ----------------------------------------------------------------------------
# 
# # Get baseline PM10 levels (latent true values)
# baseline_pm10 <- matrix(NA, nrow = n_samples, ncol = n_timepoints-1)
# for (s in 1:n_samples) {
#   for (t in 1:n_timepoints-1) {
#     baseline_pm10[s, t] <- muY_orig_array[s, t, pm10_idx]
#   }
# }
# 
# # Calculate mean PM10 across samples for each timepoint
# mean_pm10 <- colMeans(baseline_pm10)
# 
# # Categorize into quartiles
# pm10_quartiles <- cut(mean_pm10,
#                       breaks = quantile(mean_pm10, probs = c(0, 0.25, 0.5, 0.75, 1)),
#                       labels = c("Low", "Medium-Low", "Medium-High", "High"),
#                       include.lowest = TRUE)
# 
# pm10_stratified <- data.frame()
# 
# for (level in levels(pm10_quartiles)) {
#   level_idx <- which(pm10_quartiles == level)
#   
#   # Extract effects for this pollution level
#   level_abs <- delta_absolute[, level_idx, drop = FALSE]
#   level_rel <- delta_relative[, level_idx, drop = FALSE]
#   
#   # Average across timepoints within level for each sample
#   level_avg_abs_samples <- rowMeans(level_abs)
#   level_avg_rel_samples <- rowMeans(level_rel)
#   
#   pm10_stratified <- rbind(pm10_stratified, data.frame(
#     PM10_Level = level,
#     Mean_Absolute = mean(level_avg_abs_samples),
#     Lower_Absolute = quantile(level_avg_abs_samples, 0.025),
#     Upper_Absolute = quantile(level_avg_abs_samples, 0.975),
#     Mean_Relative = mean(level_avg_rel_samples),
#     Lower_Relative = quantile(level_avg_rel_samples, 0.025),
#     Upper_Relative = quantile(level_avg_rel_samples, 0.975)
#   ))
# }

# ----------------------------------------------------------------------------
# Step 8: Visualization
# ----------------------------------------------------------------------------

# Plot 1: Time series of excess deaths
p1 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_absolute, ymax = upper_absolute),
              fill = "red", alpha = 0.3) +
  geom_line(aes(y = mean_absolute), color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Excess Mortality from 1 Degree Increase in Temperature",
    subtitle = "Daily estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Excess deaths per day"
  )
ggsave("/Users/yanfei/Desktop/excess_mortality_timeseries.png", p1, width = 12, height = 5)

# Plot 2: Percent change time series
p2 <- ggplot(timepoint_summary_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_relative, ymax = upper_relative),
              fill = "blue", alpha = 0.3) +
  geom_line(aes(y = mean_relative), color = "blue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_bw() +
  labs(
    title = "Percent Change in Mortality from 1 Degree Increase in Temperature",
    subtitle = "Weekly estimates with 95% credible intervals",
    x = "Time (weeks)",
    y = "Percent change (%)"
  )
ggsave("/Users/yanfei/Desktop/percent_change_timeseries.png", p2, width = 12, height = 5)

# Plot 4: Effect by pollution level
# p4 <- ggplot(pm10_stratified, aes(x = PM10_Level, y = Mean_Relative, fill = PM10_Level)) +
#   geom_col(alpha = 0.7) +
#   geom_errorbar(aes(ymin = Lower_Relative, ymax = Upper_Relative), width = 0.2) +
#   theme_bw() +
#   labs(
#     title = "Mortality Effect by PM10 Level",
#     subtitle = "Average percent change by pollution quartile with 95% CI",
#     x = "PM10 Quartile",
#     y = "Percent change (%)"
#   ) +
#   theme(legend.position = "none")
# ggsave("/Users/yanfei/Desktop/effect_by_pollution.png", p4, width = 8, height = 5)

# Plot 5: Distribution of total excess deaths
p5 <- ggplot(data.frame(total = total_excess_samples), aes(x = total)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean(total_excess_samples), color = "red", size = 1) +
  geom_vline(xintercept = quantile(total_excess_samples, c(0.025, 0.975)),
             linetype = "dashed", color = "red", size = 0.8) +
  theme_bw() +
  labs(
    title = "Posterior Distribution: Total Excess Deaths",
    subtitle = paste0("Mean: ", round(mean(total_excess_samples), 1),
                      " (95% CI: ", round(quantile(total_excess_samples, 0.025), 1),
                      " - ", round(quantile(total_excess_samples, 0.975), 1), ")"),
    x = "Total excess deaths (258 days)",
    y = "Frequency"
  )
ggsave("/Users/yanfei/Desktop/total_excess_distribution.png", p5, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Step 9: Print formatted results for report
# ----------------------------------------------------------------------------

cat("\n", paste(rep("=", 80), collapse = ""), "\n")
cat("COUNTERFACTUAL ANALYSIS: 10% INCREASE IN TRUE PM10\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("OVERALL EFFECTS:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
cat(sprintf("Average percent change in mortality: %.2f%% (95%% CI: %.2f%%, %.2f%%)\n",
            aggregate_summary$Estimate[3],
            aggregate_summary$Lower_95[3],
            aggregate_summary$Upper_95[3]))
cat(sprintf("Average daily excess deaths: %.2f (95%% CI: %.2f, %.2f)\n",
            aggregate_summary$Estimate[2],
            aggregate_summary$Lower_95[2],
            aggregate_summary$Upper_95[2]))
cat(sprintf("Total excess deaths over study period: %.0f (95%% CI: %.0f, %.0f)\n",
            aggregate_summary$Estimate[1],
            aggregate_summary$Lower_95[1],
            aggregate_summary$Upper_95[1]))
cat(sprintf("Days with >95%% probability of positive effect: %.0f out of %d (%.1f%%)\n",
            aggregate_summary$Estimate[6], n_timepoints,
            100 * aggregate_summary$Estimate[6] / n_timepoints))

cat("\n\nEFFECT BY PM10 LEVEL (Percent change):\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
# print(pm10_stratified[, c("PM10_Level", "Mean_Relative", "Lower_Relative", "Upper_Relative")])

cat("\n\n", paste(rep("=", 80), collapse = ""), "\n")

# ----------------------------------------------------------------------------
# Step 10: Save results for reporting
# ----------------------------------------------------------------------------

# Save all results to RDS for later use
counterfactual_results <- list(
  baseline_lambda = lambda_baseline,
  counterfactual_lambda = lambda_counter,
  delta_absolute = delta_absolute,
  delta_relative = delta_relative,
  timepoint_summary = timepoint_summary_df,
  aggregate_summary = aggregate_summary,
  # seasonal_effects = seasonal_effects,
  # pm10_stratified = pm10_stratified,
  total_excess_samples = total_excess_samples,
  avg_daily_excess_samples = avg_daily_excess_samples,
  avg_percent_change_samples = avg_percent_change_samples
)

saveRDS(counterfactual_results, "/Users/yanfei/Desktop/counterfactual_results.rds")

# Also save CSV tables for easy inclusion in reports
write.csv(timepoint_summary_df, "/Users/yanfei/Desktop/daily_effects.csv", row.names = FALSE)
write.csv(aggregate_summary, "/Users/yanfei/Desktop/aggregate_effects.csv", row.names = FALSE)
# write.csv(seasonal_effects, "/Users/yanfei/Desktop/seasonal_effects.csv", row.names = FALSE)
# write.csv(pm10_stratified, "/Users/yanfei/Desktop/pm10_stratified_effects.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- /Users/yanfei/Desktop/counterfactual_results.rds\n")
cat("- /Users/yanfei/Desktop/daily_effects.csv\n")
cat("- /Users/yanfei/Desktop/aggregate_effects.csv\n")
# cat("- /Users/yanfei/Desktop/seasonal_effects.csv\n")
# cat("- /Users/yanfei/Desktop/pm10_stratified_effects.csv\n")
cat("- Various PNG plots in /Users/yanfei/Desktop/\n")
```